---
title: 'Dlx2 reprograms the transcriptome of astrocyte-derived interneurons and enforces canonical laminar positioning.'
subtitle: '1: Preprocessing 4DPI Cells'

---

```{r}
#| label: reportsetup
#| include: false
knitr::opts_chunk$set(echo = T, include = T,
                      fig.width = 20, fig.height = 10,
                      fig.align = 'center',
                      warning = F, message = F,
                      eval=T) 
```

The space below contain the codes used in this analysis for the purposes of
transparency and reproducibility:

<!-- Describe methodology -->

## Load dependencies
The code below loads all dependencies to run the script.
```{r}
#| label: load_dependencies
#| eval: true
#| include: true
#| classes: styled-output

suppressPackageStartupMessages({
    library(tidyverse)
    library(Seurat)
    library(SeuratWrappers)
    library(flexmix)
    library(ggplot2)
    library(patchwork)
   # library(DoubletFinder)
    library(scDblFinder)
    library(UpSetR)
    library(clustree)
    library(MatrixGenerics)
    library(RColorBrewer)
    library(scCustomize)
  library(dittoSeq)
})
```

## Import count matrices

For every experiment, we will prepare a samples metadata file (CSV) that should 
contain at least the following columns:

PATH : Path to gene expression matrix of each experiment

ID : Experiment name or ID that will be passed to orig.ident

The metadata file can contain any other information pertaining to the experiment 
which will be passed into the metadata dataframe of the Seurat object.

```{r}
#| label: load_metadata
#| eval: true
#| include: true
#| classes: styled-output
## Import samples meta file.
samples.meta <- read_csv("/samples.csv")

samples.meta
```

We start by importing the metadata file and create a list of Seurat objects for 
each experiment:

```{r}
#| label: load_count_matrices
#| eval: true
#| include: true
#| classes: styled-output

## Check if all paths are present.
all(dir.exists(samples.meta$Path))

## In this case I will filter for all 4dpi samples only.
samples.meta.4dpi <- samples.meta %>% 
  filter(timepoint == "4dpi")

## Creating seurat list with all samples.
seuratlist.4dpi <- apply(samples.meta.4dpi, 1, function(line){

    # Load directory containing the matrix.mtx, features.tsv, and barcodes.tsv files provided by 10X.
    sample <- Read10X(data.dir = line[["Path"]])
  
    # Create seurat objects
    sample.seurat <- CreateSeuratObject(counts = sample,
                                        project = line[["RV"]],
                                        min.cells = 4, min.features = 200)
    sample.seurat$experiment <- "RNA"
    
    # Adding additional metadata 
    #sample.seurat@meta.data <- mutate(sample.seurat@meta.data, !!!line)

    sample.seurat
})

names(seuratlist.4dpi) <- samples.meta.4dpi$RV
```

## Quantify QC metrics

Next, add more information on the quality of each cell by quantifying the 
following metrics:

- Percent mitochondria
- Label low quality cells (MiQC)
- Percent ribosomal
- Doublet annotation
- Cell cycle phase

These values will be stored in the meta.data slot of each Seurat object.

### Percent mitochondria
```{r}
#| label: percent_mito
#| eval: true
#| include: true
#| classes: styled-output

### Identify percentage of mitochondrial genes
seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
    sample$percent.mt <- PercentageFeatureSet(sample, pattern = "^mt-") 
    Idents(sample) <- "experiment"
    sample
})

# Run miQC-wrapper to model both the proportion of reads mapping to mtDNA genes 
# and the number of detected genes with mixture models in a probabilistic 
# framework to identify the low-quality cells

seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
    sample <- RunMiQC(sample, percent.mt = "percent.mt", 
                      nFeature_RNA = "nFeature_RNA", posterior.cutoff = 0.75,
                      model.slot = "flexmix_model")
    sample
})
```

### Label low quality cells (MiQC)
```{r}
#| label: MiQC_plot
#| eval: true
#| include: true
#| classes: styled-output

# Plots for MiQC
do.call(wrap_plots,lapply(seuratlist.4dpi, function(sample){
    PlotMiQC(sample, color.by = "miQC.probability") + 
      scale_color_gradient(low = "grey", high = "purple") + 
      labs(title= sample$orig.ident[1],
           subtitle=paste("Max:", round(max(sample$percent.mt[sample$miQC.probability < 0.75]), digits = 3))) +
        theme(plot.title = element_text(size = 20),
              plot.subtitle = element_text(size = 15),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size=15),
              legend.text = element_text(size=15))
}))
```

### Percent ribosomal
```{r}
#| label: percent_ribo
#| eval: true
#| include: true
#| classes: styled-output

## calculate percent ribosomal
seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
    sample$percent.rb <- PercentageFeatureSet(sample, pattern = "^Rp[sl]") 
    Idents(sample) <- "experiment"
    sample
})
```

#### Plot correlation between mitochondrial and ribosomal genes
```{r}
#| label: scatter_plot
#| eval: true
#| include: true
#| classes: styled-output

### Plot mitochondrial genes VS ribosomal genes to see correlation
do.call(wrap_plots,lapply(seuratlist.4dpi, function(sample){
    FeatureScatter(sample, feature1 = "percent.rb", feature2 = "percent.mt") +
        labs(title = sample$orig.ident[1]) +
        theme(plot.title = element_text(size = 20),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15))
}))
```

```{r}
#| label: vln_plot
#| eval: true
#| include: true
#| classes: styled-output
#| height: 8

### Plot mitochondrial genes, ribosomal genes, and number of RNA counts per dataset 
lapply(seuratlist.4dpi, function(sample){
  plots <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), function(feature){
    VlnPlot(sample, features = feature, pt.size = 0.01) +
        theme(plot.title = element_text(size = 15,face = "plain"),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15),
              legend.position="none") 
  })
  wrap_plots(plots, ncol = 4)  + 
  plot_annotation(title=sample$orig.ident[1], 
                  theme=theme(plot.title=element_text(hjust=0.5,face = "bold", size = 25)))
}) 

```
::: {.callout-warning}
## To discuss

- Further dimensionality reduction (e.g., t-SNE) and unsupervised clustering were 
also utilized in Doubletfinder's paper as pre-processing. Plus, seurat object was
pre-filtered. Removing poor quality cells.
- Homotypic doublet proportion estimation using annotation and adjusting nExp 
value using this. 
:::

```{r}
#| label: doublefinder
#| eval: true
#| include: true
#| classes: styled-output

# Increasing maximum size allowed for Global size  
options(future.globals.maxSize = 8000 * 1024^2)

## Run LogNorm, FindVariableFeatures and Scale
seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
  sample <- NormalizeData(sample, normalization.method = "LogNormalize", scale.factor = 10000)
  sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000)
  sample <- ScaleData(sample, features = rownames(sample))
})

## Run PCA
seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
    sample <- RunPCA(sample)
    sample
})
```

::: {.callout-note}
## Recommendation

We recommend using `scDblFinder` for Doublet annotation, as it has been reported
it outperforms other tools. Additionally, when plotting parameters like nCount_RNA 
and nFeature_RNA we can observe the expected behaviour in the annotated doublets.

Be aware that there is poor overlap between the outputs of `doubletFinder` and 
`scDblFinder`.
:::

```{r}
#| label: scDblfinder
#| eval: false
#| include: true
#| classes: styled-output

## annotate doublets using scDblFinder
set.seed(123)

seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
    sce <- scDblFinder(GetAssayData(sample, assay = "RNA", layer = "counts"))
    sample$scDblFinder.class <- sce$scDblFinder.class
    sample@meta.data <- sample@meta.data %>% 
      mutate(scDblFinder.class = ifelse(sample@meta.data$scDblFinder.class == "doublet", "Doublet", "Singlet"))
    sample
})
```


```{r}
#| label: scDblfinder_plot
#| eval: true
#| include: true
#| classes: styled-output

lapply(seuratlist.4dpi, function(sample){
  plots <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), function(feature){
    VlnPlot(sample, features = feature, pt.size = 0.01, group.by = "scDblFinder.class") +
        theme(plot.title = element_text(size = 15,face = "plain"),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15),
              legend.position="none") 
  })
  wrap_plots(plots, ncol = 4) + 
    plot_annotation(title=sample$orig.ident[1], 
                    theme=theme(plot.title=element_text(hjust=0.5,face = "bold", size = 25)))
}) 
```


::: {.callout-note}
## Recommendation

We recommend annotating for cell cycle phase. This is a good idea if you see 
that cell cycle drives some unwanted cluster separation purely based on these genes
:::

```{r}
#| label: cycle_genes
#| eval: true
#| include: true
#| classes: styled-output

## quantify cell cycle phase
### run cell cycle scoring
cc.genes <- cc.genes.updated.2019

##################
# To convert cc.genes to other organism of choice, run the following:
cc.genes <- lapply(cc.genes, function(genes) {
   gprofiler2::gorth(genes, 
        target_organism = "mmusculus"  
   )$ortholog_name
})
##################

seuratlist.4dpi <- lapply(seuratlist.4dpi, function(sample){
	sample <- CellCycleScoring(sample,
	                           s.features = cc.genes$s.genes,
	                           g2m.features = cc.genes$g2m.genes, 
	                           set.ident = FALSE)
})
```

## Cell cycle scoring
```{r}
#| label: cycle_genes_plot
#| eval: true
#| include: true
#| classes: styled-output

do.call(wrap_plots,lapply(seuratlist.4dpi, function(sample){
    sample@meta.data %>% 
        ggplot(aes(x=Phase, fill = Phase)) + 
        geom_bar() +
        theme_minimal() +
        scale_fill_brewer(palette = "Set2") +
        labs(title = sample$orig.ident[1]) +
        theme(plot.title = element_text(size = 20),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size=15),
              legend.text = element_text(size=15)) 
}))
```

::: {.callout-tip}
## Best practice

Another best practice recommendation is the correction of ambient RNA.
This can be done using `SoupX` or `CellBender`, the later is python-based.
`CellBender` is reported to perform best overall, however, we have not used it
before. 
:::

```{r}
#| label: save_obj
#| eval: false
#| include: true
#| classes: styled-output

# Save raw data, before filtering
saveRDS(seuratlist.4dpi, "../outputs/RDS/seuratlist_4dpi_raw.rds")
```

## Filtering poor quality cells
We recommend automatic thresholding via MAD (median absolute deviations).
We mark cells as outliers if they differ by 3 MADs. For a relatively more 
permissive filtering strategy 5 MADs can be applied.
:::

```{r}
#| label: filt_obj_MAD
#| eval: false
#| include: true
#| classes: styled-output

seuratlist.4dpi <- readRDS("/seuratlist_4dpi_raw_LogNorm.rds")

## Thresholding using median absolute deviation from the median
qc.df <- do.call(rbind,lapply(seuratlist.4dpi, function(sample){
    feature.cutoff.min <- median(sample@meta.data$nFeature_RNA) - 5*mad(sample@meta.data$nFeature_RNA)
    feature.cutoff.min <- ifelse(feature.cutoff.min < 200, 200, feature.cutoff.min)
    feature.cutoff.max <- median(sample@meta.data$nFeature_RNA) + 5*mad(sample@meta.data$nFeature_RNA)
    return(data.frame(feature.cutoff.min = feature.cutoff.min,
                      feature.cutoff.max = feature.cutoff.max))
}))

seuratlist.4dpi.filt <- lapply(seuratlist.4dpi, function(sample) {
    sample.id <- sample$orig.ident[1]
    subset(sample, 
           subset = nFeature_RNA > qc.df[sample.id,]$feature.cutoff.min & 
                    nFeature_RNA < qc.df[sample.id,]$feature.cutoff.max & 
                    miQC.keep == "keep" & 
                    scDblFinder.class == "Singlet")
})
```

## Dimensional reduction of filtered data

Performing Normalization and dimensional reduction via pca for each sample
```{r}
#| label: dim_red
#| eval: false
#| include: true
#| classes: styled-output

# Increasing maximum size allowed for Global size  
options(future.globals.maxSize = 8000 * 1024^2)

## Run LogNorm, FindVariableFeatures and Scale on filtered samples
seuratlist.4dpi.filt <- lapply(seuratlist.4dpi.filt, function(sample){
  sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000)
  sample <- ScaleData(sample, features = rownames(sample))
})

## Perform dimensional reduction
seuratlist.4dpi.filt <- lapply(seuratlist.4dpi.filt, function(sample){
	sample <- RunPCA(sample)
  sample 
})

saveRDS(seuratlist.4dpi.filt, "/seuratlist_4dpi_filt_LogNorm.rds")
```

## Annotation of reporter genes

We observe the expression of reporter genes per sample
```{r}
#| label: reporter_genes_exp
#| eval: true
#| include: true
#| classes: styled-output

# Plot expression of reporter genes
genes_of_interest <- c("Mscarlet3LTR", "GFPU3", "Mscarlet", "EGFPtetO", "NeonGreen")

lapply(seuratlist.4dpi.filt, function(sample){
  FeaturePlot(sample, features = genes_of_interest, reduction = "pca")
})
```

We proceed to annotate cells per sample according to the expression of 
reporter genes.
```{r}
#| label: reporter_genes_anno
#| eval: false
#| include: true
#| classes: styled-output

# Annotate each sample based on expression of reporter genes
seuratlist.4dpi.filt <- lapply(seuratlist.4dpi.filt, function(sample) {
  for (gene in genes_of_interest) {
    if (gene %in% rownames(sample)) {
      sample@meta.data[[paste0(gene, ".pos")]] <- ifelse(
        FetchData(object = sample, vars = c(gene), layer = "counts") > 0, "Pos", "Neg"
      )
    } 
  }
  return(sample)
})

#Saving sample names
names.seuratlist.4dpi.filt <- names(seuratlist.4dpi.filt)

# Annotate filtering based on expression of reporter genes
seuratlist.4dpi.filt <- lapply(names(seuratlist.4dpi.filt), function(sample_name) {
  seurat_object <- seuratlist.4dpi.filt[[sample_name]]
  
  # Add the new column based on conditions using if-else
  if (sample_name == "Control") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  } else if (sample_name == "Ascl1WT") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" & seurat_object@meta.data$GFPU3.pos == "Neg"
  } else if (sample_name == "Ascl1WTBcl2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" & seurat_object@meta.data$GFPU3.pos == "Pos"
  } else if (sample_name == "Ascl1SA6") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" & seurat_object@meta.data$GFPU3.pos == "Neg"
  } else if (sample_name == "Ascl1SA6Bcl2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" & seurat_object@meta.data$GFPU3.pos == "Pos"
  } else if (sample_name == "Ascl1SA62") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet.pos == "Pos" & seurat_object@meta.data$EGFPtetO.pos == "Neg"
  } else if (sample_name == "Ascl1SA6Dlx2") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$EGFPtetO.pos == "Pos"
  } else if (sample_name == "Ascl1WTSA6") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet.pos == "Pos" & seurat_object@meta.data$NeonGreen.pos == "Pos"
  } else if (sample_name == "Ascl1WTDlx2") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$NeonGreen.pos == "Pos"
  } else if (sample_name == "Dlx2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet.pos == "Neg" & seurat_object@meta.data$NeonGreen.pos == "Pos"
  } else if (sample_name == "Ascl1SA6Dlx2Bcl2") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  }else if (sample_name == "Ascl1SA6Dlx2_2") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$GFPU3.pos == "Neg" & seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  }
  return(seurat_object)
})

# Renaming list 
names(seuratlist.4dpi.filt) <- names.seuratlist.4dpi.filt
```

This table reports the number of cells annotated according to the expression
of reporter genes/combination of genes.
```{r}
#| label: reporter_genes_tally
#| eval: true
#| include: true
#| classes: styled-output

## Number of cells per sample after filtering and annotated based on reporter genes
cell.tally <- do.call(rbind,lapply(seuratlist.4dpi.filt, function(sample){
  tibble(samples = unique(sample$orig.ident),
         total = nrow(sample@meta.data),
         transcript.keep = nrow(filter(sample@meta.data, transcript.filtering.keep == TRUE)))
}))

cell.tally
```

The following plots show sparse to no expression of expected reporter genes in 
cells for certain samples.

```{r}
#| label: reporter_genes_plot
#| eval: true
#| include: true
#| classes: styled-output
#| fig-width: 20
#| fig.height: 10

# Plotting annotation
do.call(wrap_plots,lapply(seuratlist.4dpi.filt, function(sample){
  DimPlot(sample, group.by = "transcript.filtering.keep", reduction = "pca")  +
        labs(title = sample$orig.ident[1])
}))
```

In the interest of retaining cells affected by this, we implemented a method
that considers the clustering of cells and how far away each cell is from the 
centroid (Mahalanobis distance) to determine outliers. Additionally, 
those cells that express the expected combination of reporter genes (true cells)
are retained as well.

```{r}
#| label: Clustering
#| eval: false
#| include: true
#| classes: styled-output
 
# Clustering of at low resolution for further filtering
seuratlist.4dpi.filt <- lapply(seuratlist.4dpi.filt, function(sample){
	sample <- FindNeighbors(sample, dims = 1:50)
  sample <- FindClusters(sample, resolution = 0.1)
  sample
})
```

```{r}
#| label: Clustering_plot
#| eval: true
#| include: true
#| classes: styled-output
#| fig-width: 20
#| fig.height: 10

# Plotting clusters
do.call(wrap_plots,lapply(seuratlist.4dpi.filt, function(sample){
  DimPlot(sample, reduction = "pca") +
        labs(title = sample$orig.ident[1])
}))
```

```{r}
#| label: Clustering_tally
#| eval: true
#| include: true
#| classes: styled-output

## Number of cells per cluster per sample after filtering and annotated based on reporter genes
cluster.tally <- do.call(rbind,lapply(seuratlist.4dpi.filt, function(sample){
  sample@meta.data %>% 
    group_by(seurat_clusters,transcript.filtering.keep) %>% 
    tally() %>% 
    mutate(samples = unique(sample$orig.ident)) %>% 
    relocate(samples, .before = seurat_clusters)
}))

cluster.tally
```

```{r}
#| label: outlier_anno
#| eval: false
#| include: true
#| classes: styled-output

## Annotating outliers for each cluster
seuratlist.4dpi.filt <- lapply(seuratlist.4dpi.filt, function(sample){
  
  # Number of principal components to use
  PC <- 10
  
  #Extract PCA embeddings
  pca_df <- as.data.frame(Embeddings(sample, reduction = "pca"))[,1:PC]
  
  # Add cluster annotations
  pca_df$cluster <- sample$seurat_clusters
  pca_df$keep.anno <- sample$transcript.filtering.keep
  
  # Calculate Mahalanobis distance per cluster using lapply
  results <- do.call(rbind, lapply(unique(pca_df$cluster), function(cl) {
    
    # Subset data for the cluster
    cluster_data <- pca_df %>% filter(cluster == cl)
    
    # Calculate mean and covariance using only TRUE cells  
    dat.center <- colMeans(cluster_data[, 1:PC]) ### This is not using only TRUE cells
    #dat.center <- colMeans(cluster_data[cluster_data$keep.anno, 1:PC]) ### This would
    dat.cov <- cov(cluster_data[, 1:PC])
    #dat.cov <- cov(cluster_data[cluster_data$keep.anno, 1:PC]) ### Same here
    
    # Compute Mahalanobis distance
    cluster_data$mdist <- mahalanobis(
      x = cluster_data[, 1:PC],
      center = dat.center,
      cov = dat.cov
    )
    
    # Calculate cut-off with Chi-squared (90% confidence)
    cutoff <- qchisq(p = 0.90, df = PC)  
    
    # Add outliers based on Mahalanobis distance
    cluster_data <- cluster_data %>%
      mutate(Outlier = ifelse(mdist > cutoff, "Yes", "No"))
    
    # Return cluster results as a list
    return(cluster_data)
  }))
  
  # Keep all true cells based on reporter genes
  results <- results %>% 
    mutate(Outlier = ifelse(Outlier == "Yes" & keep.anno == TRUE, "No", Outlier)) %>% 
    rownames_to_column("cell") %>% 
    select(c("cell","Outlier", "mdist"))

  # Append annotation to meta data
  sample@meta.data <- sample@meta.data %>% 
     rownames_to_column("cell") %>% 
     left_join(results, by = join_by("cell")) %>% 
      column_to_rownames("cell")
   
  return(sample)
})
```

The following plots show the cells annotated as outliers and the percent of cell 
that would be retained after filtering out said outliers.
```{r}
#| label: Outlier_plot
#| eval: true
#| include: true
#| classes: styled-output

# Plot distribution of outliers
lapply(seuratlist.4dpi.filt, function(sample){
  
  # Calculate number of outliers per sample
  outlier_table <- table(sample$Outlier)

  # Calculate the percentage of retained cells 
  percentage <- (outlier_table["No"] / sum(outlier_table)) * 100
  
  # Plot outlier cells per cluster in PCA
  DimPlot(sample, reduction = "pca", group.by = "seurat_clusters", split.by = "Outlier", shuffle = T) +
    labs(title = unique(sample$orig.ident),
         subtitle = paste("% retained -", round(percentage,digits = 2)),
         x = "PC1", y = "PC2")
})
```

These are the final number of cells retained
```{r}
#| label: Outlier_tally
#| eval: true
#| include: true
#| classes: styled-output

## Number of cells per sample after filtering for outliers
Non_outlier.tally <- do.call(rbind,lapply(seuratlist.4dpi.filt, function(sample){
  sample@meta.data %>% 
    group_by(Outlier) %>% 
    tally() %>% 
    mutate(samples = unique(sample$orig.ident)) %>% 
    filter(Outlier == "No") %>% 
    relocate(samples, .before = Outlier)
}))

Non_outlier.tally
```

```{r}
#| label: Outlier_filter
#| eval: false
#| include: true
#| classes: styled-output

## Filtering data to remove outliers
seuratlist.4dpi.filtered <- lapply(seuratlist.4dpi.filt, function(sample) {
    sample.id <- sample$orig.ident[1]
    subset(sample, 
           subset = Outlier == "No")
})
```

## Dimensional reduction of filtered data

Performing Normalization and dimensional reduction via pca for each sample
```{r}
#| label: dim_red_pca
#| eval: false
#| include: true
#| classes: styled-output

# Increasing maximum size allowed for Global size  
options(future.globals.maxSize = 8000 * 1024^2)

## Run LogNorm, FindVariableFeatures and Scale on filtered samples
seuratlist.4dpi.filtered <- lapply(seuratlist.4dpi.filtered, function(sample){
  sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000)
  sample <- ScaleData(sample, features = rownames(sample))
})

## Perform dimensional reduction
seuratlist.4dpi.filtered <- lapply(seuratlist.4dpi.filtered, function(sample){
	sample <- RunPCA(sample)
  sample 
})
```

```{r}
#| label: saveRDS
#| eval: false
#| include: true
#| classes: styled-output

# Save filtered object
saveRDS(seuratlist.4dpi.filtered, "/ss_list_filt_reporter_genes_LogNorm.rds")
```

```{r}
#| label: pca_plot
#| eval: true
#| include: true
#| classes: styled-output
#| fig-width: 20
#| fig.height: 10

do.call(wrap_plots,lapply(seuratlist.4dpi.filtered, function(sample){
  DimPlot(sample, reduction = "pca") +
        labs(title = sample$orig.ident[1])
}))
```

We obtain the overlap of variable features per sample. In this case, 40.6% of 
3000 variable features are shared between all samples.

```{r}
#| label: varfeat_overlap
#| eval: true
#| include: true
#| classes: styled-output

# Computing the overlap of variable features between samples
var.feat <- lapply(seuratlist.4dpi.filtered, VariableFeatures)

upset(fromList(var.feat), sets = names(var.feat),
      sets.bar.color = "#56B4E9", order.by = "freq",main.bar.color = "#660099")
```

```{r}
seuratlist.4dpi.filtered <- readRDS("/ss_list_filt_reporter_genes_LogNorm.rds")
```


# Integration of samples
```{r}
#| label: harmony_integration
#| eval: false
#| include: true
#| classes: styled-output


## Merge Samples
merged.seurat <- merge(x = seuratlist.4dpi.filtered[[1]],
                       y = seuratlist.4dpi.filtered[2:length(seuratlist.4dpi.filtered)],
                       merge.data = TRUE, add.cell.ids = c("Control", "Ascl1WT", "Ascl1WTBcl2", "Ascl1SA6", "Ascl1SA6Bcl2","Ascl1SA62", "Ascl1SA6Dlx2", "Ascl1WTSA6", 
                                                           "Ascl1WTDlx2", "Dlx2","Ascl1SA6Dlx2_2", "Ascl1SA6Dlx2Bcl2"))

merged.seurat <- merged.seurat %>% 
  ScaleData(features = rownames(merged.seurat)) %>%
  RunPCA()


saveRDS(merged.seurat, "/seurat_4dpi_merged_LogNorm.rds")
```



```{r}
mergeAll_filtered <- readRDS('/seurat_4dpi_merged_LogNorm.rds')
```

```{r}
# Update RV values
mergeAll_filtered$RV <- ifelse(mergeAll_filtered$RV == "Ascl1SA6Dlx2_2", "SA6Dlx2", mergeAll_filtered$RV)
unique(mergeAll_filtered$RV)

# Update Cell IDs

mergeAll_filtered@meta.data <- mergeAll_filtered@meta.data %>%
  rownames_to_column(var = "old.names") %>%
  mutate(new.names = gsub("^Ascl1SA6Dlx2_2_", "SA6Dlx2_", old.names)) %>%
  column_to_rownames(var = "old.names")

mergeAll_filtered <- RenameCells(mergeAll_filtered, new.names = mergeAll_filtered$new.names)        
         
grep("^SA6Dlx2", rownames(mergeAll_filtered@meta.data), value = TRUE)
grep("Ascl1SA6Dlx2_2", rownames(mergeAll_filtered@meta.data), value = TRUE)

# Sample Name Mapping
sample_name_mapping <- c(
  "Control" = "A4",
  "Ascl1WT" = "A5",
  "Ascl1WTBcl2" = "A6",
  "Ascl1SA6" = "A7",
  "Ascl1SA6Bcl2" = "A8",
  "Ascl1SA62" = "A9",
  "Ascl1SA6Dlx2" = "A10",
  "Ascl1WTSA6" = "A11",
  "Ascl1WTDlx2" = "A12",
  "Dlx2" = "A13",
  "SA6Dlx2" = "A17",
  "Ascl1SA6Dlx2Bcl2" = "A18"
)

# Copy the 'RV' column to 'sample_name' and replace values based on the mapping
mergeAll_filtered$sample_name <- recode(mergeAll_filtered$RV, !!!sample_name_mapping)

# Rename cells
mergeAll_filtered$sample <- rownames(mergeAll_filtered@meta.data)
mergeAll_filtered@meta.data <- separate(mergeAll_filtered@meta.data, col = 'sample', into = c('Sample','Barcode'), sep = '_')

# Generate the 'cell_name_loom' column
mergeAll_filtered@meta.data <- mergeAll_filtered@meta.data %>%
  mutate(cell_name_loom = glue("{sample_name}_cellranger_output:{Barcode}")) %>%
  mutate(cell_name_loom = str_replace_all(cell_name_loom, "-\\d+", "x")) %>%
  mutate(cell_name_loom = glue("{cell_name_loom}-{sample_name}"))

mergeAll_filtered <- RenameCells(mergeAll_filtered, new.names = mergeAll_filtered@meta.data$cell_name_loom)

```

```{r}
mergeAll_filtered <- readRDS("/seurat_4dpi_merged_LogNorm_newID.rds") 
```



###continue with new Cell IDS 

```{r}
mergeAll_filtered <- readRDS("/seurat_4dpi_merged_LogNorm_newID.rds")             
Idents(mergeAll_filtered) <- "RV"
mergeAll_filtered <- subset(mergeAll_filtered, downsample=3000)
print(table(mergeAll_filtered$RV))
saveRDS(mergeAll_filtered, '/seurat_4dpi_merged_LogNorm_newID_3000.rds')
```



```{r install.data, eval=FALSE, include=FALSE}
mergeAll_filtered <- readRDS("/seurat_4dpi_merged_LogNorm_newID_3000.rds")
```

```{r preprocessing7, error=TRUE, message=FALSE, include=FALSE}
mergeAll_filtered <- FindNeighbors(object = mergeAll_filtered, dims = 1:35) 
mergeAll_filtered <- FindClusters(object = mergeAll_filtered, resolution = 0.1)
mergeAll_filtered <- RunUMAP(object = mergeAll_filtered, dims = 1:40)
```

```{r plotting all clusters, error=TRUE}
Idents(mergeAll_filtered) <- "seurat_clusters"
DimPlot(mergeAll_filtered, reduction = "umap", group.by = 'RV', pt.size = 0.5) + Blank_Theme()
DimPlot(mergeAll_filtered, reduction = "umap", group.by = 'orig.ident', pt.size = 0.5) + Blank_Theme()
DimPlot(mergeAll_filtered, reduction = "umap", group.by = 'seurat_clusters', pt.size = 0.5) + Blank_Theme()
DimPlot(mergeAll_filtered, reduction = "umap", label = TRUE, pt.size = 0.5) 
```

```{r}
# Violin plot of number of genes and library count for each sample
VlnPlot(mergeAll_filtered, features = c("nFeature_RNA", "nCount_RNA"), group.by = "seurat_clusters", pt.size = 0)
```
## remove low transcript count cells 
```{r}
mergeAll_filtered <- subset(mergeAll_filtered, subset = seurat_clusters %in% c(4, 9, 10), invert = TRUE)
DimPlot(mergeAll_filtered, group.by = "seurat_clusters", shuffle = TRUE)
```

```{r install.data, eval=FALSE, include=FALSE}
saveRDS(mergeAll_filtered, '/seurat_4dpi_merged_LogNorm_newID_3000_removedlowcounts.rds')
```


```{r install.data, eval=FALSE, include=FALSE}
mergeAll_filtered <- readRDS("/seurat_4dpi_merged_LogNorm_newID_3000_removedlowcounts.rds")
```













