---
title: 'Dlx2 reprograms the transcriptome of astrocyte-derived interneurons and enforces canonical laminar positioning.'
subtitle: '1: Preprocessing 2 and 4wPI Cells'

---

```{r}
#| label: reportsetup
#| include: false
knitr::opts_chunk$set(echo = T, include = T,
                      fig.width = 20, fig.height = 10,
                      fig.align = 'center',
                      warning = F, message = F,
                      eval=T) 
```

The space below contain the codes used in this analysis for the purposes of
transparency and reproducibility:

<!-- Describe methodology -->

## Load dependencies
The code below loads all dependencies to run the script.
```{r}
#| label: load_dependencies
#| eval: true
#| include: true
#| classes: styled-output

suppressPackageStartupMessages({
    library(tidyverse)
    library(Seurat)
    library(SeuratWrappers)
    library(flexmix)
    library(ggplot2)
    library(patchwork)
    library(DoubletFinder)
    library(scDblFinder)
  library(tidyverse)
    library(Seurat)
    library(SeuratWrappers)
    library(flexmix)
    library(ggplot2)
    library(patchwork)
    library(DoubletFinder)
    library(scDblFinder)
    library(UpSetR)
    library(clustree)
    library(MatrixGenerics)
    library(RColorBrewer)
    library(scCustomize)
})
```

## Import count matrices

For every experiment, we will prepare a samples metadata file (CSV) that should 
contain at least the following columns:

PATH : Path to gene expression matrix of each experiment

ID : Experiment name or ID that will be passed to orig.ident

The metadata file can contain any other information pertaining to the experiment 
which will be passed into the metadata dataframe of the Seurat object.

```{r}
#| label: load_metadata
#| eval: true
#| include: true
#| classes: styled-output
## Import samples meta file.
samples.meta <- read_csv("/samples.csv")

samples.meta
```

We start by importing the metadata file and create a list of Seurat objects for 
each experiment:

```{r}
#| label: load_count_matrices
#| eval: false
#| include: true
#| classes: styled-output

#all(dir.exists(samples.meta$Path))

## In this case I will filter for all 2wpi samples only.
samples.meta.2wpi <- samples.meta %>% 
  filter(timepoint == "2wpi")

## Creating seurat list with all samples.
seuratlist.2wpi <- apply(samples.meta.2wpi, 1, function(line){

    # Load directory containing the matrix.mtx, features.tsv, and barcodes.tsv files provided by 10X.
    sample <- Read10X(data.dir = line[["Path"]])
  
      # Create seurat objects
      sample.seurat <- CreateSeuratObject(counts = sample,
                                          project = line[["RV"]],
                                          min.cells = 4, min.features = 200)
      sample.seurat$experiment <- "RNA"
      # Adding additional metadata 
      sample.seurat@meta.data <- mutate(sample.seurat@meta.data, !!!line)
  
      sample.seurat
})

names(seuratlist.2wpi) <- samples.meta.2wpi$RV
```

## Quantify QC metrics

Next, add more information on the quality of each cell by quantifying the 
following metrics:

- Percent mitochondria
- Label low quality cells (MiQC)
- Percent ribosomal
- Doublet annotation
- Cell cycle phase

These values will be stored in the meta.data slot of each Seurat object.

### Percent mitochondria
```{r}
#| label: percent_mito
#| eval: false
#| include: true
#| classes: styled-output

### Identify percentage of mitochondrial genes
seuratlist.2wpi <- lapply(seuratlist.2wpi, function(sample){
    sample$percent.mt <- PercentageFeatureSet(sample, pattern = "^mt-") 
    Idents(sample) <- "experiment"
    sample
})

# Run miQC-wrapper to model both the proportion of reads mapping to mtDNA genes 
# and the number of detected genes with mixture models in a probabilistic 
# framework to identify the low-quality cells

seuratlist.2wpi <- lapply(seuratlist.2wpi, function(sample){
    sample <- RunMiQC(sample, percent.mt = "percent.mt", 
                      nFeature_RNA = "nFeature_RNA", posterior.cutoff = 0.75,
                      model.slot = "flexmix_model")
    sample
})
```

### Orig Label low quality cells (MiQC)
```{r, fig.width=16, fig.height=8}
#| label: MiQC_plot
#| eval: true
#| include: true
#| classes: styled-output

# Plots for MiQC
do.call(wrap_plots,lapply(seuratlist.2wpi, function(sample){
    PlotMiQC(sample, color.by = "miQC.probability") + 
      scale_color_gradient(low = "grey", high = "purple") + 
      labs(title= sample$orig.ident[1],
           subtitle=paste("Max:", round(max(sample$percent.mt[sample$miQC.probability < 0.75]), digits = 3))) +
        theme(plot.title = element_text(size = 20),
              plot.subtitle = element_text(size = 15),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size=15),
              legend.text = element_text(size=15))
}))
```

### Modified Label low quality cells (MiQC)
```{r, fig.width=16, fig.height=8}
#| label: MiQC_plot_2
#| eval: true
#| include: true
#| classes: styled-output

plots <- lapply(seuratlist.2wpi, function(sample) {
  # Debugging: Check sample structure
  tryCatch({
    if (length(sample$percent.mt) > 0 && length(sample$miQC.probability) > 0) {
      # Check if filtering leads to empty data
      valid_indices <- which(sample$miQC.probability < 0.75)
      
      if (length(valid_indices) > 0) {
        max_mt <- round(max(sample$percent.mt[valid_indices]), digits = 3)
      } else {
        max_mt <- NA
      }
      
      # Convert orig.ident to character
      title <- as.character(sample$orig.ident[1])
      
      # Generate plot
      PlotMiQC(sample, color.by = "miQC.probability") + 
        scale_color_gradient(low = "grey", high = "purple") + 
        labs(title = title,
             subtitle = paste("Max:", max_mt)) +
        theme(plot.title = element_text(size = 20),
              plot.subtitle = element_text(size = 15),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15))
    } else {
      NULL
    }
  }, error = function(e) {
    message("Error in sample: ", as.character(sample$orig.ident[1]), " - ", e$message)
    NULL
  })
})

# Remove NULL results
plots <- Filter(Negate(is.null), plots)

# Combine plots
if (length(plots) > 0) {
  do.call(wrap_plots, plots)
} else {
  message("No valid plots to display.")
}
```



### Percent ribosomal
```{r}
#| label: percent_ribo
#| eval: false
#| include: true
#| classes: styled-output

## calculate percent ribosomal
seuratlist.2wpi <- lapply(seuratlist.2wpi, function(sample){
    sample$percent.rb <- PercentageFeatureSet(sample, pattern = "^Rp[sl]") 
    Idents(sample) <- "experiment"
    sample
})
```

#### Plot correlation between mitochondrial and ribosomal genes
```{r, fig.width=16, fig.height=8}
#| label: scatter_plot
#| eval: true
#| include: true
#| classes: styled-output

### Plot mitochondrial genes VS ribosomal genes to see correlation
do.call(wrap_plots,lapply(seuratlist.2wpi, function(sample){
    FeatureScatter(sample, feature1 = "percent.rb", feature2 = "percent.mt") +
        labs(title = sample$orig.ident[1]) +
        theme(plot.title = element_text(size = 20),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15))
}))
```

```{r}
#| label: vln_plot
#| eval: true
#| include: true
#| classes: styled-output
#| height: 8

### Plot mitochondrial genes, ribosomal genes, and number of RNA counts per dataset 
lapply(seuratlist.2wpi, function(sample){
  plots <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), function(feature){
    VlnPlot(sample, features = feature, pt.size = 0.01) +
        theme(plot.title = element_text(size = 15,face = "plain"),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15),
              legend.position="none") 
  })
  wrap_plots(plots, ncol = 4)  + 
  plot_annotation(title=sample$orig.ident[1], 
                  theme=theme(plot.title=element_text(hjust=0.5,face = "bold", size = 25)))
}) 

```

```{r}
#| label: scDblfinder
#| eval: false
#| include: true
#| classes: styled-output

## annotate doublets using scDblFinder
set.seed(123)

seuratlist.2wpi <- lapply(seuratlist.2wpi, function(sample){
    sce <- scDblFinder(GetAssayData(sample, assay = "RNA", layer = "counts"))
    sample$scDblFinder.class <- sce$scDblFinder.class
    sample@meta.data <- sample@meta.data %>% 
      mutate(scDblFinder.class = ifelse(sample@meta.data$scDblFinder.class == "doublet", "Doublet", "Singlet"))
    sample
})
```

####original 
```{r}
#| label: scDblfinder_plot
#| eval: true
#| include: true
#| classes: styled-output

scdbl.plot <- lapply(seuratlist.2wpi, function(sample){
    VlnPlot(sample, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), group.by = "scDblFinder.class", ncol = 4) + 
    theme(legend.position = 'right') +
    labs(fill = "scDblFinder")
})

lapply(names(seuratlist.2wpi), function(sample){
  scdbl.plot[[sample]]  + 
  plot_annotation(title=sample, 
                  theme=theme(plot.title=element_text(hjust=0.5,face = "bold",size = 20)))
})
```


```{r}
#| label: cycle_genes
#| eval: false
#| include: true
#| classes: styled-output

## quantify cell cycle phase
### run cell cycle scoring
cc.genes <- cc.genes.updated.2019

##################
# To convert cc.genes to other organism of choice, run the following:
cc.genes <- lapply(cc.genes, function(genes) {
   gprofiler2::gorth(genes, 
        target_organism = "mmusculus"  
   )$ortholog_name
})
##################

seuratlist.2wpi <- lapply(seuratlist.2wpi, function(sample){
	sample <- NormalizeData(sample)
    CellCycleScoring(sample, 
                     s.features = cc.genes$s.genes, 
                     g2m.features = cc.genes$g2m.genes, 
                     set.ident = FALSE)
})
```

## Cell cycle scoring
```{r, fig.width=16, fig.height=8}
#| label: cycle_genes_plot
#| eval: true
#| include: true
#| classes: styled-output

do.call(wrap_plots,lapply(seuratlist.2wpi, function(sample){
    sample@meta.data %>% 
        ggplot(aes(x=Phase, fill = Phase)) + 
        geom_bar() +
        theme_minimal() +
        scale_fill_brewer(palette = "Set2") +
        labs(title = sample$orig.ident[1]) +
        theme(plot.title = element_text(size = 20),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size=15),
              legend.text = element_text(size=15)) 
}))
```


```{r}
#| label: save_obj
#| eval: false
#| include: true
#| classes: styled-output

# Save raw data, before filtering
saveRDS(seuratlist.2wpi, "/seuratlist_2and4wpi_raw.rds")

seuratlist.2wpi <- readRDS("/seuratlist_2and4wpi_raw.rds")
```

## Filtering data

We filter our data usingthresholding via 5 MAD (median absolute deviations). Mito cutoff at 1%, why miQC.keep??
```{r}
#| label: filt_obj_MAD
#| eval: false
#| include: true
#| classes: styled-output

## Thresholding using median absolute deviation from the median
qc.df <- do.call(rbind,lapply(seuratlist.2wpi, function(sample){
    feature.cutoff.min <- median(sample@meta.data$nFeature_RNA) - 5*mad(sample@meta.data$nFeature_RNA)
    feature.cutoff.min <- ifelse(feature.cutoff.min < 100, 100, feature.cutoff.min)
    feature.cutoff.max <- median(sample@meta.data$nFeature_RNA) + 5*mad(sample@meta.data$nFeature_RNA)
    return(data.frame(feature.cutoff.min = feature.cutoff.min,
                      feature.cutoff.max = feature.cutoff.max))
}))

seuratlist.2wpi.filt <- lapply(seuratlist.2wpi, function(sample) {
    sample.id <- sample$orig.ident[1]
    subset(sample, 
           subset = nFeature_RNA > qc.df[sample.id,]$feature.cutoff.min & 
                    nFeature_RNA < qc.df[sample.id,]$feature.cutoff.max & 
                    miQC.keep == "keep" & 
                    scDblFinder.class == "Singlet")
})

```

## Dimensional reduction of filtered data

Performing Normalization and dimensional reduction via pca for each sample
```{r}
#| label: dim_red
#| eval: false
#| include: true
#| classes: styled-output

# Increasing maximum size allowed for Global size  
options(future.globals.maxSize = 8000 * 1024^2)

## Run LogNorm, FindVariableFeatures and Scale on filtered samples
seuratlist.2wpi.filt <- lapply(seuratlist.2wpi.filt, function(sample){
  sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000)
  sample <- ScaleData(sample, features = rownames(sample))
})

## Perform dimensional reduction
seuratlist.2wpi.filt <- lapply(seuratlist.2wpi.filt, function(sample){
	sample <- RunPCA(sample)
  sample 
})

saveRDS(seuratlist.2wpi.filt, "/seuratlist_2and4wpi_LogNorm_filtered.rds")
```


```{r}
#| label: vln_plot_postfilt
#| eval: true
#| include: true
#| classes: styled-output
#| height: 8

### Plot mitochondrial genes, ribosomal genes, and number of RNA counts per dataset 
lapply(seuratlist.2wpi.filt, function(sample){
  plots <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), function(feature){
    VlnPlot(sample, features = feature, pt.size = 0.01) +
        theme(plot.title = element_text(size = 15,face = "plain"),
              axis.title.x = element_text(size = 15),
              axis.title.y = element_text(size = 15),
              axis.text = element_text(size = 15),
              legend.key.size = unit(0.5, 'cm'), 
              legend.key.height = unit(0.5, 'cm'), 
              legend.key.width = unit(0.5, 'cm'), 
              legend.title = element_text(size = 15),
              legend.text = element_text(size = 15),
              legend.position="none") 
  })
  wrap_plots(plots, ncol = 4)  + 
  plot_annotation(title=sample$orig.ident[1], 
                  theme=theme(plot.title=element_text(hjust=0.5,face = "bold", size = 25)))
}) 

```


```{r}
# Tools' versions used during this analysis
sessionInfo()
```






