---
title: 'BC021: Deconstructing in vivo glia-to-neuron conversion in the postnatal mouse cortex.'
subtitle: '2: Retroviral transcript annotation of samples and sampling'
date: last-modified
author: 
    - name : Karla Lozano Gonzalez
      email : karla.lozano_gonzalez@kcl.ac.uk
      orcid : 0000-0003-0329-4668
      affiliation: CDN Bioinformatics Core
      affiliation-url: https://emckclac.sharepoint.com/sites/IoPPNcdnb/SitePages/Bioinformatics.aspx
    - name : Fursham Hamid
      email : fursham.hamid@kcl.ac.uk
      url : https://twitter.com/FurshamH
      orcid : 0000-0002-6846-7382
      affiliation: CDN Bioinformatics Core
      affiliation-url: https://devneuro.org/cdn/index.php
    
format:
    html:
        page-layout: full
        theme: pulse
        highlight: espresso
        css: /home/cdn-bc/Github_repo/fuRsh/inst/resources/styles.css
        smooth-scroll: true
        self-contained: true
        toc-depth: 3
        toc: true
        toc-location: right
        toc-title: Contents
        code-fold: false
        df-print: paged
        grid:
            sidebar-width: 150px
            body-width: 950px
            margin-width: 150px
            gutter-width: 1em
fig-cap-location: margin
tbl-cap-location: margin
fig-align: center
title-block-banner: /home/cdn-bc/Github_repo/fuRsh/man/figures/banner.jpg
title-blocks-style: plain
filters:
   - /home/cdn-bc/Github_repo/fuRsh/inst/resources/include-code-files.lua
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: reportsetup
#| include: false
knitr::opts_chunk$set(echo = T, include = T,
                      fig.width = 8, fig.height = 6,
                      fig.align = 'center',
                      warning = F, message = F,
                      eval=T) 
```


The space below contain the codes used in this analysis for the purposes of
transparency and reproducibility:

<!-- Describe methodology -->

## Load dependencies
The code below loads all dependencies to run the script.
```{r}
#| label: load_dependencies
#| eval: true
#| include: true
#| classes: styled-output

suppressPackageStartupMessages({
    library(tidyverse)
    library(Seurat)
    library(ggplot2)
    library(patchwork)
})
```

```{r}
#| label: read_obj
#| eval: true
#| include: false
#| classes: styled-output

seuratlist.2wpi.filt <- readRDS("/Users/alexiscooper/The Francis Crick Dropbox/GuillemotF/Dropbox Alexis Cooper/Cooper_et_al_2025/2-4wpi/preprocessing_2_4wpi_Nuclei_LogNorm/output/seuratlist_2and4wpi_LogNorm_filtered.rds")
```


## Annotation of reporter genes

We observe the expression of reporter genes per sample
```{r}
#| label: reporter_genes_exp
#| eval: true
#| include: true
#| classes: styled-output

# Plot expression of reporter genes
genes_of_interest <- c("Mscarlet3LTR", "GFPU3")

lapply(seuratlist.2wpi.filt, function(sample){
  FeaturePlot(sample, features = genes_of_interest, reduction = "pca")
})
```

We proceed to annotate cells per sample according to the expression of 
reporter genes.

```{r}
lapply(seuratlist.2wpi.filt, function(x) {
  colnames(x@meta.data)
})
```

```{r}
#| label: reporter_genes_anno
#| eval: false
#| include: true
#| classes: styled-output

# Annotate each sample based on expression of reporter genes
seuratlist.2wpi.filt <- lapply(seuratlist.2wpi.filt, function(sample) {
  for (gene in genes_of_interest) {
    if (gene %in% rownames(sample)) {
      sample@meta.data[[paste0(gene, ".pos")]] <- ifelse(
        FetchData(object = sample, vars = c(gene), layer = "counts") > 0, "Pos", "Neg"
      )
    } 
  }
  return(sample)
})

#Saving sample names
names.seuratlist.2wpi.filt <- names(seuratlist.2wpi.filt)

# Annotate filtering based on expression of reporter genes
seuratlist.2wpi.filt <- lapply(names(seuratlist.2wpi.filt), function(sample_name) {
  seurat_object <- seuratlist.2wpi.filt[[sample_name]]

  # Add the new column based on conditions using if-else
  if (sample_name == "Control") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  } else if (sample_name == "Ascl1WT") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" 
  } else if (sample_name == "Ascl1WTBcl2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" 
  } else if (sample_name == "Ascl1SA6") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" 
  } else if (sample_name == "Ascl1SA6Bcl2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" 
  } else if (sample_name == "Bcl2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Neg" & seurat_object@meta.data$GFPU3 == "Pos"
  } else if (sample_name == "Ascl1SA6Dlx2") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  } else if (sample_name == "Ascl1SA6Dlx2_2") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  } else if (sample_name == "Ascl1WTSA6") {
    seurat_object@meta.data$transcript.filtering.keep <- seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  } else if (sample_name == "Ascl1WTDlx2") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$Mscarlet3LTR.pos == "Pos" 
  } else if (sample_name == "Ascl1SA6Bcl2_mat") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  }else if (sample_name == "Ascl1SA6Dlx2_mat") {
    seurat_object@meta.data$transcript.filtering.keep <-  seurat_object@meta.data$Mscarlet3LTR.pos == "Pos"
  }
  return(seurat_object)
})

# Renaming list 
names(seuratlist.2wpi.filt) <- names.seuratlist.2wpi.filt
```

This table reports the number of cells annotated according to the expression
of reporter genes/combination of genes.
```{r}
#| label: reporter_genes_tally
#| eval: true
#| include: true
#| classes: styled-output

## Number of cells per sample after filtering and annotated based on reporter genes
cell.tally <- do.call(rbind,lapply(seuratlist.2wpi.filt, function(sample){
  tibble(samples = unique(sample$orig.ident),
    total = nrow(sample@meta.data),
       transcript.keep = nrow(filter(sample@meta.data, transcript.filtering.keep == TRUE)))
}))

cell.tally
```


The following plots show sparse to no expression of expected reporter genes in 
cells for certain samples.

```{r}
#| label: reporter_genes_plot
#| eval: true
#| include: true
#| classes: styled-output
#| fig-width: 20
#| fig.height: 10

# Plotting annotation
do.call(wrap_plots,lapply(seuratlist.2wpi.filt, function(sample){
  DimPlot(sample, group.by = "transcript.filtering.keep", reduction = "pca")  +
        labs(title = sample$orig.ident[1])
}))
```

In the interest of retaining cells affected by this, we implemented a method
that considers the clustering of cells and how far away each cell is from the 
centroid (Mahalanobis distance) to determine outliers. Additionally, 
those cells that express the expected combination of reporter genes (true cells)
are retained as well.

```{r}
#| label: Clustering
#| eval: false
#| include: true
#| classes: styled-output
 
# Clustering of at low resolution for further filtering
seuratlist.2wpi.filt <- lapply(seuratlist.2wpi.filt, function(sample){
	sample <- FindNeighbors(sample, dims = 1:50)
  sample <- FindClusters(sample, resolution = 0.1)
  sample
})
```

```{r}
#| label: Clustering_plot
#| eval: true
#| include: true
#| classes: styled-output
#| fig-width: 20
#| fig.height: 10

# Plotting clusters
do.call(wrap_plots,lapply(seuratlist.2wpi.filt, function(sample){
  DimPlot(sample, reduction = "pca") +
        labs(title = sample$orig.ident[1])
}))
```

```{r}
#| label: Clustering_tally
#| eval: true
#| include: true
#| classes: styled-output

## Number of cells per cluster per sample after filtering and annotated based on reporter genes
cluster.tally <- do.call(rbind,lapply(seuratlist.2wpi.filt, function(sample){
  sample@meta.data %>% 
    group_by(seurat_clusters,transcript.filtering.keep) %>% 
    tally() %>% 
    mutate(samples = unique(sample$orig.ident)) %>% 
    relocate(samples, .before = seurat_clusters)
}))

cluster.tally
```

```{r}
#| label: outlier_anno
#| eval: false
#| include: true
#| classes: styled-output

## Annotating outliers for each cluster
seuratlist.2wpi.filt <- lapply(seuratlist.2wpi.filt, function(sample){
  
  # Number of principal components to use
  PC <- 10
  
  #Extract PCA embeddings
  pca_df <- as.data.frame(Embeddings(sample, reduction = "pca"))[,1:PC]
  
  # Add cluster annotations
  pca_df$cluster <- sample$seurat_clusters
  pca_df$keep.anno <- sample$transcript.filtering.keep
  
  # Calculate Mahalanobis distance per cluster using lapply
  results <- do.call(rbind, lapply(unique(pca_df$cluster), function(cl) {
    
    # Subset data for the cluster
    cluster_data <- pca_df %>% filter(cluster == cl)
    
    # Calculate mean and covariance using only TRUE cells
    dat.center <- colMeans(cluster_data[, 1:PC])
    dat.cov <- cov(cluster_data[, 1:PC])
    
    # Compute Mahalanobis distance
    cluster_data$mdist <- mahalanobis(
      x = cluster_data[, 1:PC],
      center = dat.center,
      cov = dat.cov
    )
    
    # Calculate cut-off with Chi-squared (90% confidence)
    cutoff <- qchisq(p = 0.90, df = PC)  
    
    # Add outliers based on Mahalanobis distance
    cluster_data <- cluster_data %>%
      mutate(Outlier = ifelse(mdist > cutoff, "Yes", "No"))
    
    # Return cluster results as a list
    return(cluster_data)
  }))
  
  # Keep all true cells based on reporter genes
  results <- results %>% 
    mutate(Outlier = ifelse(Outlier == "Yes" & keep.anno == TRUE, "No",Outlier)) %>% 
    rownames_to_column("cell") %>% 
    select(c("cell","Outlier", "mdist"))

  # Append annotation to meta data
  sample@meta.data <- sample@meta.data %>% 
     rownames_to_column("cell") %>% 
     left_join(results, by = join_by("cell")) %>% 
      column_to_rownames("cell")
   
  return(sample)
})
```

The following plots show the cells annotated as outliers and the percent of cell 
that would be retained after filtering out said outliers.
```{r}
#| label: Outlier_plot
#| eval: true
#| include: true
#| classes: styled-output

# Plot distribution of outliers
lapply(seuratlist.2wpi.filt, function(sample){
  
  # Calculate number of outliers per sample
  outlier_table <- table(sample$Outlier)

  # Calculate the percentage of retained cells 
  percentage <- (outlier_table["No"] / sum(outlier_table)) * 100
  
  # Plot outlier cells per cluster in PCA
  DimPlot(sample, reduction = "pca", group.by = "seurat_clusters", split.by = "Outlier", shuffle = T) +
    labs(title = unique(sample$orig.ident),
         subtitle = paste("% retained -", round(percentage,digits = 2)),
         x = "PC1", y = "PC2")
})
```

These are the final number of cells retained
```{r}
#| label: Outlier_tally
#| eval: true
#| include: true
#| classes: styled-output

## Number of cells per sample after filtering for outliers
Non_outlier.tally <- do.call(rbind,lapply(seuratlist.2wpi.filt, function(sample){
  sample@meta.data %>% 
    group_by(Outlier) %>% 
    tally() %>% 
    mutate(samples = unique(sample$orig.ident)) %>% 
    filter(Outlier == "No") %>% 
    relocate(samples, .before = Outlier)
}))

Non_outlier.tally
```


```{r}
#| label: Outlier_filter
#| eval: false
#| include: true
#| classes: styled-output

## Filtering data to remove outliers
seuratlist.2wpi.filtered <- lapply(seuratlist.2wpi.filt, function(sample) {
    sample.id <- sample$orig.ident[1]
    subset(sample, 
           subset = Outlier == "No")
})
```

## Dimensional reduction of filtered data

Performing Normalization and dimensional reduction via pca for each sample
```{r}
#| label: dim_red_pca
#| eval: false
#| include: true
#| classes: styled-output

# Increasing maximum size allowed for Global size  
options(future.globals.maxSize = 8000 * 1024^2)

## Run LogNorm, FindVariableFeatures and Scale on filtered samples
seuratlist.2wpi.filtered <- lapply(seuratlist.2wpi.filtered, function(sample){
  sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000)
  sample <- ScaleData(sample, features = rownames(sample))
})

## Perform dimensional reduction
seuratlist.2wpi.filtered <- lapply(seuratlist.2wpi.filtered, function(sample){
	sample <- RunPCA(sample)
  sample 
})
```


```{r}

###make sure all samples dont have more than 1% mito
# Filter out cells with more than 1% mitochondrial genes
seuratlist.2wpi.filtered <- lapply(seuratlist.2wpi.filtered, function(sample) {
  # Filter cells based on mitochondrial percentage
  sample <- subset(sample, subset = percent.mt < 1)
  return(sample)
})

```


```{r}
#| label: saveRDS
#| eval: false
#| include: true
#| classes: styled-output

# Save filtered object
saveRDS(seuratlist.2wpi.filtered, "/Users/alexiscooper/The Francis Crick Dropbox/GuillemotF/Dropbox Alexis Cooper/Cooper_et_al_2025/2-4wpi/preprocessing_2_4wpi_Nuclei_LogNorm/output/seuratlist_2wpi_LogNorm_filt_reporter_genes.rds")
```

```{r}
seuratlist.2wpi.filtered <- readRDS("/Users/alexiscooper/The Francis Crick Dropbox/GuillemotF/Dropbox Alexis Cooper/Cooper_et_al_2025/2-4wpi/preprocessing_2_4wpi_Nuclei_LogNorm/output/seuratlist_2wpi_LogNorm_filt_reporter_genes.rds")
```

```{r}
#| label: session_info
#| eval: true
#| include: true
#| classes: styled-output

# Tools' versions used during this analysis
sessionInfo()
```

