
```{r}
#| label: reportsetup
#| include: false
knitr::opts_chunk$set(echo = T, include = T,
                      fig.width = 8, fig.height = 6,
                      fig.align = 'center',
                      warning = F, message = F,
                      eval=T) 
```


The space below contain the codes used in this analysis for the purposes of
transparency and reproducibility:

<!-- Describe methodology -->

## Load dependencies
The code below loads all dependencies to run the script.
```{r}
#| label: load_dependencies
#| eval: true
#| include: true
#| classes: styled-output

suppressPackageStartupMessages({
    library(tidyverse)
    library(Seurat)
    library(ggplot2)
    library(patchwork)
    library(UpSetR)
    library(harmony)
    library(clustree)
    library(MatrixGenerics)
  library(glmGamPoi)
})
```

# Loading of data 

```{r}
#| label: load_data
#| eval: false
#| include: true
#| classes: styled-output

seuratlist.2wpi.filtered <- readRDS("/seuratlist_2wpi_LogNorm_filt_reporter_genes.rds")
```

```{r}
#| label: vln_plot_postfilt
#| eval: true
#| include: true
#| classes: styled-output
#| height: 8

# Plot mitochondrial genes, ribosomal genes, and number of RNA counts per dataset
lapply(seuratlist.2wpi.filtered, function(sample) {
  plots <- lapply(c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.rb"), function(feature) {
    VlnPlot(sample, features = feature, pt.size = 0.01) +  # No group.by to avoid clustering split
      theme(plot.title = element_text(size = 15, face = "plain"),
            axis.title.x = element_text(size = 15),
            axis.title.y = element_text(size = 15),
            axis.text = element_text(size = 15),
            legend.key.size = unit(0.5, 'cm'),
            legend.key.height = unit(0.5, 'cm'),
            legend.key.width = unit(0.5, 'cm'),
            legend.title = element_text(size = 15),
            legend.text = element_text(size = 15),
            legend.position = "none")
  })
  
  # Combine and arrange the plots
  wrap_plots(plots, ncol = 4) + 
    plot_annotation(title = sample$orig.ident[1], 
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 25)))
})

```

#Performing Normalization and dimensional reduction via pca for each sample
```{r}
#| label: dim_red_pca
#| eval: false
#| include: true
#| classes: styled-output

# Increasing maximum size allowed for Global size  
options(future.globals.maxSize = 8000 * 1024^2)

## Run LogNorm, FindVariableFeatures and Scale on filtered samples
seuratlist.2wpi.filtered <- lapply(seuratlist.2wpi.filtered, function(sample){
  sample <- FindVariableFeatures(sample, selection.method = "vst", nfeatures = 2000)
  sample <- ScaleData(sample, features = rownames(sample))
})

## Perform dimensional reduction
seuratlist.2wpi.filtered <- lapply(seuratlist.2wpi.filtered, function(sample){
	sample <- RunPCA(sample)
  sample 
})
```
```{r}
saveRDS(seuratlist.2wpi.filtered, "/seuratlist_2wpi_LogNorm_filt_reporter_genes.rds")
```

```{r}
#####Scaling
old_total_counts_Ascl1SA6Bcl2_mat <- colSums(GetAssayData(seuratlist.2wpi.filtered[["Ascl1SA6Bcl2_mat"]], assay = "RNA", layer = "counts"))
mean(old_total_counts_Ascl1SA6Bcl2_mat)

# Step 1: Get the counts for Ascl1SA6Bcl2_mat and Ascl1SA6Dlx2_mat using GetAssayData
counts_Ascl1SA6Bcl2_mat <- GetAssayData(seuratlist.2wpi.filtered[["Ascl1SA6Bcl2_mat"]], assay = "RNA", layer = "counts")
counts_Ascl1SA6Dlx2_mat <- GetAssayData(seuratlist.2wpi.filtered[["Ascl1SA6Dlx2_mat"]], assay = "RNA", layer = "counts")

# Step 2: Calculate total counts per cell for both samples
total_counts_Ascl1SA6Bcl2_mat <- colSums(counts_Ascl1SA6Bcl2_mat)
total_counts_Ascl1SA6Dlx2_mat <- colSums(counts_Ascl1SA6Dlx2_mat)

# Step 3: Calculate the mean total counts per cell for both samples
mean_counts_Ascl1SA6Bcl2_mat <- mean(total_counts_Ascl1SA6Bcl2_mat)
mean_counts_Ascl1SA6Dlx2_mat <- mean(total_counts_Ascl1SA6Dlx2_mat)

# Step 4: Calculate scaling factor
scaling_factor <- mean_counts_Ascl1SA6Dlx2_mat / mean_counts_Ascl1SA6Bcl2_mat  ###scaling_factor 0.4938886
scaling_factor
```

```{r}
#| label: varfeat_overlap
#| eval: false
#| include: true
#| classes: styled-output
# Assign each Seurat object to a new variable without the 'seurat_' prefix
Ascl1SA6 <- seuratlist.2wpi.filtered[["Ascl1SA6"]]
Bcl2 <- seuratlist.2wpi.filtered[["Bcl2"]]
Ascl1SA6Bcl2 <- seuratlist.2wpi.filtered[["Ascl1SA6Bcl2"]]
Ascl1SA6Dlx2 <- seuratlist.2wpi.filtered[["Ascl1SA6Dlx2"]]
Ascl1WTSA6 <- seuratlist.2wpi.filtered[["Ascl1WTSA6"]]
Ascl1WT <- seuratlist.2wpi.filtered[["Ascl1WT"]]
Ascl1WTBcl2 <- seuratlist.2wpi.filtered[["Ascl1WTBcl2"]]
Ascl1WTDlx2 <- seuratlist.2wpi.filtered[["Ascl1WTDlx2"]]
Ascl1SA6Dlx2_2 <- seuratlist.2wpi.filtered[["Ascl1SA6Dlx2_2"]]
Control <- seuratlist.2wpi.filtered[["Control"]]
Ascl1SA6Bcl2_mat <- seuratlist.2wpi.filtered[["Ascl1SA6Bcl2_mat"]]
Ascl1SA6Dlx2_mat <- seuratlist.2wpi.filtered[["Ascl1SA6Dlx2_mat"]]

# Set seed for reproducibility
set.seed(123)
# Downsample Control to 1000 cells
Control_cells <- sample(Cells(Control), size = 2000, replace = FALSE)
Control <- subset(Control, cells = Control_cells)

# Define the scaling factor
scaling_factor <- 0.4938886
scaling_factor <- 0.5
# Extract raw counts for Ascl1SA6Bcl2_mat
counts_Ascl1SA6Bcl2_mat <- GetAssayData(Ascl1SA6Bcl2_mat, assay = "RNA", slot = "counts")

# Apply the scaling factor to adjust the counts
scaled_counts_Ascl1SA6Bcl2_mat <- counts_Ascl1SA6Bcl2_mat * 0.5 #scaling_factor 0.6653017

# Update the counts in the Ascl1SA6Bcl2_mat Seurat object
Ascl1SA6Bcl2_mat[["RNA"]] <- SetAssayData(
  object = Ascl1SA6Bcl2_mat[["RNA"]],
  slot = "counts",
  new.data = scaled_counts_Ascl1SA6Bcl2_mat
)

# Recalculate nCount_RNA for Ascl1SA6Bcl2_mat
Ascl1SA6Bcl2_mat <- AddMetaData(
  object = Ascl1SA6Bcl2_mat,
  metadata = colSums(GetAssayData(Ascl1SA6Bcl2_mat, assay = "RNA", slot = "counts")), col.name = "nCount_RNA")


mean_nCount_Ascl1SA6Bcl2_mat <- mean(Ascl1SA6Bcl2_mat@meta.data$nCount_RNA)

print(paste("Mean nCount_RNA for Ascl1SA6Bcl2_mat:", mean_nCount_Ascl1SA6Bcl2_mat))

mean_nCount_Ascl1SA6Dlx2_mat <- mean(Ascl1SA6Dlx2_mat@meta.data$nCount_RNA)

print(paste("Mean nCount_RNA for Ascl1SA6Dlx2_mat:", mean_nCount_Ascl1SA6Dlx2_mat))



# Merge all Seurat objects
mergeAll <- merge(
  x = Ascl1SA6,
  y = list(Ascl1SA6Bcl2,Ascl1SA6Dlx2,Ascl1WTSA6,Ascl1WT, Ascl1WTBcl2,Ascl1WTDlx2,Ascl1SA6Dlx2_2,Control,Ascl1SA6Bcl2_mat,Ascl1SA6Dlx2_mat),
  add.cell.ids = c(
    "Ascl1SA6","Ascl1SA6Bcl2", "Ascl1SA6Dlx2","Ascl1WTSA6","Ascl1WT", "Ascl1WTBcl2", "Ascl1WTDlx2","Ascl1SA6Dlx2_2", "Control", "Ascl1SA6Bcl2_mat", "Ascl1SA6Dlx2_mat"
  ),
  project = "MergedProject",
  merge.data = TRUE  # Merge normalized data as well
)
remove( "Ascl1SA6","Ascl1SA6Bcl2", "Ascl1SA6Dlx2","Ascl1WTSA6","Ascl1WT", "Ascl1WTBcl2", "Ascl1WTDlx2","Ascl1SA6Dlx2_2", "Control", "Ascl1SA6Bcl2_mat", "Ascl1SA6Dlx2_mat", "counts_Ascl1SA6Bcl2_mat", "counts_Ascl1SA6Dlx2_mat")
```


```{r}
saveRDS(seuratlist.2wpi.filtered, "/seuratlist_2wpi_LogNorm_filt_reporter_genes_scale_SA6Blc2_mat.rds")
```




We obtain the overlap of variable features per sample. In this case, 40.6% of 
3000 variable features are shared between all samples.

```{r}
#| label: varfeat_overlap
#| eval: true
#| include: true
#| classes: styled-output

# Computing the overlap of variable features between samples
var.feat <- lapply(seuratlist.2wpi.filtered, VariableFeatures)

upset(fromList(var.feat), sets = names(var.feat),
      sets.bar.color = "#56B4E9", order.by = "freq",main.bar.color = "#660099")
```

```{r}
# Ensure data layers are joined before continuing
mergeAll[["RNA"]] <- JoinLayers(mergeAll[["RNA"]])

# Recalculate nCount_RNA for merged object after scaling
mergeAll <- AddMetaData(mergeAll, metadata = colSums(GetAssayData(mergeAll, assay = "RNA", slot = "counts")), col.name = "nCount_RNA")

# Normalize the data using LogNormalize
DefaultAssay(mergeAll) <- "RNA"
mergeAll <- NormalizeData(mergeAll, normalization.method = "LogNormalize", scale.factor = 10000)

# Identify variable features
mergeAll <- FindVariableFeatures(mergeAll, selection.method = "vst", nfeatures = 3000)

# Scale the data using the variable features
mergeAll <- ScaleData(mergeAll, features = VariableFeatures(mergeAll))

# Run PCA using the variable features
mergeAll <- RunPCA(mergeAll, npcs = 30)

# Run downstream analysis (neighbors, clustering, UMAP)
mergeAll <- FindNeighbors(mergeAll, dims = 1:20)
mergeAll <- FindClusters(mergeAll, resolution = 0.5)
mergeAll <- RunUMAP(mergeAll, dims = 1:23) #23

# Visualize the results for all RVs (groups)
DimPlot(mergeAll, reduction = "umap", group.by = "RV") + scale_x_reverse() + scale_y_reverse() # Group by RV to ensure all groups are shown
```
FeaturePlot(mergeAll, c("P2ry12", "Aldh1l1","Lamp5", "Sst", "Pvalb"), max.cutoff = 1.5,reduction = "umap")


```{r}
###order samples 
mergeAll$RV <- factor(mergeAll$RV,levels=c("Control","Bcl2" ,"Ascl1WT", "Ascl1WTBcl2", "Ascl1SA6", "Ascl1SA6Bcl2", "Ascl1SA6Dlx2", "Ascl1WTSA6", "Ascl1WTDlx2", "Ascl1SA6Dlx2_2", "Ascl1SA6Bcl2_mat", "Ascl1SA6Dlx2_mat"))
```


```{r}
saveRDS(mergeAll, "/2wpi_merged_logNorm.rds")
```

