---
title: "CGE figure"
output: html_document
date: "2026-02-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
library(clusterProfiler)
library(cowplot)
library(dittoSeq)
library(dplyr)
library(forcats)
library(ggplot2)
library(gridExtra)
library(magrittr)
library(Nebulosa)
library(org.Mm.eg.db)
library(patchwork)
library(pheatmap)
library(purrr)
library(qs)
library(RColorBrewer)
library(scCustomize)
library(Seurat)
library(SeuratData)
library(SeuratDisk)
library(tibble)
library(tidyr)
library(tidyverse)
library(viridis)
library(ComplexHeatmap)
})
```


## STEP 1 ##Figure 7b/c and Figure S14 b
```{r}
# -------------------------------
# Subset only P28 interneurons
# -------------------------------
Neurons_P28 <- subset(Neurons, subset = grepl("P28$", sample) & !is.na(sample))
Idents(Neurons_P28) <- "sample"

# -------------------------------
# Define MGE and CGE cell groups
# -------------------------------
MGE <- WhichCells(Neurons_P28, idents = c("Pvalb_P28", "Sst_P28"))
CGE <- WhichCells(Neurons_P28, idents = c("Vip_P28", "Lamp5_P28"))

# -------------------------------
# Create a new metadata column for MGE/CGE identity
# -------------------------------
Neurons_P28$MGE_CGE <- NA
Neurons_P28$MGE_CGE[MGE] <- "MGE"
Neurons_P28$MGE_CGE[CGE] <- "CGE"

# -------------------------------
# Set new identities
# -------------------------------
Idents(Neurons_P28) <- "MGE_CGE"

# -------------------------------
# Find DEGs between MGE and CGE interneurons
# -------------------------------
MGE_vs_CGE_DEG <- FindMarkers(
  Neurons_P28,
  ident.1 = "MGE",
  ident.2 = "CGE",
  logfc.threshold = 0.25,
  min.pct = 0.25
)

# Add direction of upregulation
MGE_vs_CGE_DEG$up_in <- ifelse(MGE_vs_CGE_DEG$avg_log2FC > 0, "Up_in_MGE", "Up_in_CGE")
list_TFs <- read.csv(  "/Users/alexiscooper/Library/CloudStorage/Dropbox-TheFrancisCrick/GuillemotF/Dropbox Alexis Cooper/Cooper_et_al_2025/MGE:CGE/list_TFs.csv", header = TRUE, stringsAsFactors = FALSE)


### extract TF from list 
MGE_vs_CGE_TFs <- MGE_vs_CGE_DEG[rownames(MGE_vs_CGE_DEG) %in% list_TFs$TFs,]

Neurons <- AddModuleScore(Neurons, features = rownames(MGE_vs_CGE_TFs)[MGE_vs_CGE_TFs$up_in == "Up_in_CGE"], ctrl = 15, name = "CGE")
Neurons <- AddModuleScore(Neurons, features = rownames(MGE_vs_CGE_TFs)[MGE_vs_CGE_TFs$up_in == "Up_in_MGE"], ctrl = 15, name = "MGE")
Neurons <- subset(Neurons, subset = !is.na(sample))

# Define color palette (same as your existing scheme)
colors <- c(
  "Ascl1SA6Bcl2_2wpi" = "#9400D3",
  "Ascl1SA6Bcl2_4wpi" = "#6A5ACD",
  "Ascl1SA6Dlx2_2wpi" = "#32CD32",
  "Ascl1SA6Dlx2_4wpi" = "#00FA9A",
  "Pvalb_P10" = "#DFA9A9",
  "Pvalb_P28" = "#A88BB3",
  "Sst_P10" = "#D9D3A0",
  "Sst_P28" = "#C9B08A",
  "Vip_P10" = "#9BB199",
  "Vip_P28" = "#D0F0C0",
  "Lamp5_P10" = "#ADD8E6",
  "Lamp5_P28" = "#8BAAC0"
)

Idents(Neurons) <- "sample"

# Define a reusable boxplot function
bp <- function(f) {
  SCpubr::do_BoxPlot(Neurons, feature = f, split.by = "sample") +
    scale_fill_manual(values = colors) +
   # coord_cartesian(ylim = c(-0.5, 1.5)) +
    ggtitle(f)
}

# 📊 Plot CGE and MGE module scores side by side
bp("CGE1") | bp("MGE1")

rownames(MGE_vs_CGE_TFs)[MGE_vs_CGE_TFs$up_in == "Up_in_CGE"]
rownames(MGE_vs_CGE_TFs)[MGE_vs_CGE_TFs$up_in == "Up_in_MGE"]
```

####STEP 2
```{r}
library(Seurat)
library(SCpubr)
library(RColorBrewer)
library(dplyr)
library(patchwork)

# ====== Subset relevant samples ======
plot_samples <- c( "Pvalb_P28", "Sst_P28", "Vip_P28", "Lamp5_P28",   "Ascl1SA6Bcl2_4wpi", "Ascl1SA6Dlx2_4wpi")
Neurons_subset <- subset(Neurons, subset = sample %in% plot_samples)
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD", "Ascl1SA6Dlx2_4wpi"="#00FA9A", "Pvalb_P28"="#666666", "Sst_P28"="#9C9C9C", "Vip_P28"="#D7D7D7", "Lamp5_P28"="#E6E6E6")
bp <- function(f) SCpubr::do_BoxPlot(Neurons_subset, feature = f, split.by = "sample") +
  scale_fill_manual(values = colors) +
  coord_cartesian(ylim = c(-2, 3)) +
  ggtitle(f)

# ====== Plot MGE, CGE, and combined (MGE − CGE) scores ======
Neurons_subset$CGE_minus_MGE <- Neurons_subset$CGE1 - Neurons_subset$MGE1

 bp("CGE1") | bp("MGE1") | bp("CGE_minus_MGE")

```


### this for FIgure pooling CGE and MGE!!!! STEP 3
```{r}
# ====== Pooled groups and subset ======
Neurons_subset$pooled_group <- ifelse(Neurons_subset$sample %in% c("Pvalb_P28","Sst_P28"), "MGE",
                              ifelse(Neurons_subset$sample %in% c("Vip_P28","Lamp5_P28"), "CGE",
                              Neurons_subset$sample))
Neurons_subset <- subset(Neurons_subset, subset = pooled_group %in% c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","CGE","MGE"))
Neurons_subset$pooled_group <- factor(Neurons_subset$pooled_group, levels = c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","CGE","MGE"))
Idents(Neurons_subset) <- "pooled_group"

# ====== Colors ======
colors_pooled <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A","CGE"="#5F5F5F","MGE"="#BDBDBD")

# ====== Boxplot function ======

bp <- function(f) SCpubr::do_BoxPlot(Neurons_subset, feature=f) + scale_fill_manual(values=colors_pooled) + coord_cartesian(ylim=c(-2,3)) + ggtitle(f) + theme(text=element_text(size=16))


# ====== Combined score ======
Neurons_subset$CGE_minus_MGE <- Neurons_subset$CGE1 - Neurons_subset$MGE1

# ====== Plot ======
bp("CGE1") | bp("MGE1")

```

### 4 rank TF avg_log2FC ensures TFs are truly enriched.

#(pct.1 - pct.2) rewards TFs that are highly expressed in your group and minimally in the other.

# Sorting by this score gives you robust and specific TFs first, which is exactly what you want for CGE vs MGE.
```{r}
MGE_vs_CGE_TFs$score <- MGE_vs_CGE_TFs$avg_log2FC * (MGE_vs_CGE_TFs$pct.1 - MGE_vs_CGE_TFs$pct.2)

# CGE-specific TFs
CGE_TFs <- MGE_vs_CGE_TFs %>%
  filter(up_in=="Up_in_CGE") %>%
  arrange(desc(score))

cat("Number of CGE TFs:", nrow(CGE_TFs), "\n")
print(rownames(CGE_TFs))

# MGE-specific TFs
MGE_TFs <- MGE_vs_CGE_TFs %>%
  filter(up_in=="Up_in_MGE") %>%
  arrange(desc(score))

cat("Number of MGE TFs:", nrow(MGE_TFs), "\n")
print(rownames(MGE_TFs))

```

## STEP 5 last step for plotting 
```{r}
library(Seurat); library(SCpubr); library(dplyr)

# ---- Top 10 CGE then MGE TFs ----
top10_TFs <- intersect(c(head(rownames(CGE_TFs),10), head(rownames(MGE_TFs),10)), rownames(Neurons_subset))

# ---- Stacked violin plot ----
VlnPlot(Neurons_subset, features=top10_TFs, group.by="pooled_group", fill.by="ident",
        stack=TRUE, flip=TRUE, pt.size=0, cols=c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A","CGE"="#5F5F5F","MGE"="#BDBDBD")) +
  NoLegend() + ggtitle("Top 10 TFs per Group (CGE → MGE)") + ylab("Gene Name") +
  theme(panel.grid=element_blank(), panel.border=element_rect(color="black", fill=NA), text=element_text(size=14))

# ---- Top 10 TFs per group ----
top10_CGE <- intersect(head(rownames(CGE_TFs),10), rownames(Neurons_subset))
top10_MGE <- intersect(head(rownames(MGE_TFs),10), rownames(Neurons_subset))
cols <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A","CGE"="#5F5F5F","MGE"="#BDBDBD")

# ---- Create individual plots ----
p_CGE <- VlnPlot(Neurons_subset, features=top10_CGE, group.by="pooled_group", fill.by="ident",
                 stack=TRUE, flip=TRUE, pt.size=0, cols=cols) +
  NoLegend() + ggtitle("Top 10 CGE TFs") + ylab("Gene Name") +
  theme(panel.grid=element_blank(), panel.border=element_rect(color="black", fill=NA), text=element_text(size=14))

p_MGE <- VlnPlot(Neurons_subset, features=top10_MGE, group.by="pooled_group", fill.by="ident",
                 stack=TRUE, flip=TRUE, pt.size=0, cols=cols) +
  NoLegend() + ggtitle("Top 10 MGE TFs") + ylab("Gene Name") +
  theme(panel.grid=element_blank(), panel.border=element_rect(color="black", fill=NA), text=element_text(size=14))

# ---- Combine side by side ----
p_CGE | p_MGE

```


### NOw for Seperate TFs 

## this works STEP 1
```{r}
library(Seurat); library(SCpubr); library(dplyr); library(patchwork)

# ---- P28 DEGs ----
Neurons_P28 <- subset(Neurons, grepl("P28$", sample)); Idents(Neurons_P28) <- "sample"

library(MAST)

DEG_P28 <- FindAllMarkers(interneuron_cells, test.use="MAST", logfc.threshold=0.25, min.pct=0.25, only.pos=TRUE)
#DEG_P28 <- FindAllMarkers(Neurons_P28, logfc.threshold=.25, min.pct=.25, only.pos=TRUE)

# ---- TF filter ----
list_TFs <- read.csv("/Users/alexiscooper/Library/CloudStorage/Dropbox-TheFrancisCrick/GuillemotF/Dropbox Alexis Cooper/Cooper_et_al_2025/MGE:CGE/list_TFs.csv")
DEG_TFs <- DEG_P28[DEG_P28$gene %in% list_TFs$TFs, ]

# ---- Robust TF selection per subtype ----
subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")
top_TFs <- do.call(rbind, lapply(subtypes, function(cl){
  tmp <- DEG_TFs[DEG_TFs$cluster==cl & DEG_TFs$pct.1>=.35 & DEG_TFs$pct.2<=.6 & DEG_TFs$avg_log2FC>=.3, ]
  if(nrow(tmp)==0) return(NULL)
  tmp$score <- tmp$avg_log2FC * (tmp$pct.1 - tmp$pct.2)
  tmp[order(-tmp$score), ]
}))

# ---- Module scores (SAFE) ----
DefaultAssay(Neurons) <- "RNA"; genes <- rownames(Neurons)
for(cl in subtypes){
  g <- intersect(top_TFs$gene[top_TFs$cluster==cl], genes)
  if(length(g)>=3) Neurons <- AddModuleScore(Neurons, list(g), ctrl=15, name=paste0(cl,"_TF"))
}

# ---- Plot subset ----
samples_levels <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Vip_P28","Lamp5_P28","Pvalb_P28","Sst_P28")
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A",
            "Vip_P28"="#5F5F5F","Lamp5_P28"="#4D4D4D","Pvalb_P28"="#BDBDBD","Sst_P28"="#9E9E9E")

Neurons_plot <- subset(Neurons, sample %in% samples_levels)
Neurons_plot$sample <- factor(Neurons_plot$sample, levels=samples_levels)
Idents(Neurons_plot) <- "sample"
bp <- function(f) SCpubr::do_BoxPlot(Neurons_plot, feature=f, split.by="sample") +
  scale_fill_manual(values=colors) + coord_cartesian(ylim=c(-1,1)) + ggtitle(f)
# ---- Plots ----
bp("Lamp5_P28_TF1") | bp("Vip_P28_TF1") | bp("Sst_P28_TF1") | bp("Pvalb_P28_TF1")
# ---- Inspect TFs ----
lapply(subtypes, function(cl) top_TFs$gene[top_TFs$cluster==cl])

```

```{r}
# ---- Sample order and colors ----
samples_levels <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")
my_colors <- c(
  "Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A",
  "Lamp5_P28"="#4D4D4D","Vip_P28"="#5F5F5F","Sst_P28"="#9E9E9E","Pvalb_P28"="#BDBDBD"
)
Neurons_plot$sample <- factor(Neurons_plot$sample, levels=samples_levels)
Idents(Neurons_plot) <- "sample"

# ---- Top 5 TFs per subtype from top_TFs ----
top5_TFs <- unlist(lapply(subtypes, function(cl){
  tfs <- top_TFs[top_TFs$cluster==cl, ]
  head(intersect(tfs$gene[order(-tfs$avg_log2FC)], rownames(Neurons_plot)), 5)
}))

# ---- Stacked violin plot ----
VlnPlot(Neurons_plot, features=top5_TFs, group.by="sample", fill.by="ident",
        stack=TRUE, flip=TRUE, pt.size=0, cols=my_colors) +
  NoLegend() + ggtitle("Top 5 TFs per Subtype") +
  ylab("Gene Name") +
  theme(panel.grid=element_blank(), panel.border=element_rect(color="black", fill=NA))

```



```{r}
library(Seurat); library(SCpubr); library(patchwork)

# ---- Colors & plotting order ----
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A","Lamp5_P28"="#4D4D4D","Vip_P28"="#5F5F5F","Sst_P28"="#9E9E9E","Pvalb_P28"="#BDBDBD")
samples_levels <- names(colors)
subtypes <- c("Lamp5_P28","Vip_P28","Pvalb_P28", "Sst_P28"); ascl1_samples <- samples_levels[1:4]

Merged_Neurons <- subset(Neurons, subset = sample %in% c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Lamp5_P28","Vip_P28","Pvalb_P28", "Sst_P28"))

Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels=samples_levels); Idents(Merged_Neurons) <- "sample"
Merged_Neurons[["RNA"]] <- JoinLayers(Merged_Neurons[["RNA"]])

P28_Neurons <- subset(Neurons, subset = sample %in% c("Lamp5_P28","Vip_P28","Pvalb_P28", "Sst_P28"))

library(MAST)

DEG_P28_Neurons <- FindAllMarkers(P28_Neurons, test.use="MAST",
                           logfc.threshold=0.25, min.pct=0.25, only.pos=TRUE)
# ---- Load TF list ----
list_TFs <- read.csv("/Users/alexiscooper/Library/CloudStorage/Dropbox-TheFrancisCrick/GuillemotF/Dropbox Alexis Cooper/Cooper_et_al_2025/MGE:CGE/list_TFs.csv",
                     header=TRUE, stringsAsFactors=FALSE)

# ---- Keep only TF DEGs ----
DEG_P28_Neurons_TFs <- DEG_P28_Neurons[DEG_P28_Neurons$gene %in% list_TFs$TFs, ]

# ---- Step 1: Filter TFs per subtype (expression ≥20%, optional fold-change >0) ----
subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")  # exact names

top_TFs_list <- lapply(subtypes, function(cl) {
  tmp <- DEG_P28_Neurons_TFs[
    DEG_P28_Neurons_TFs$cluster == cl &
    DEG_P28_Neurons_TFs$pct.1 >= 0.35 &
    DEG_P28_Neurons_TFs$pct.2 <= 0.6 &
    DEG_P28_Neurons_TFs$avg_log2FC >= 0.3, 
  ]
  if(nrow(tmp) == 0) return(NULL)
  tmp$enrichment <- tmp$pct.1 / tmp$pct.2
  tmp[order(-tmp$enrichment, -tmp$avg_log2FC), ]
})

top_TFs <- do.call(rbind, top_TFs_list)



 # top 10 by score → you’re computing a combined score of fold-change × fraction of expressing cells. This sorts TFs from highest to lowest “score” and picks the top 10. But it does not remove TFs below a certain fold-change — a TF with a low avg_log2FC but high pct.1 could still appear.


# ---- Add module scores to Merged_Neurons ----
DefaultAssay(Merged_Neurons) <- "RNA"
for(sub in subtypes){
  genes <- top_TFs$gene[top_TFs$cluster==sub]
  if(length(genes)>0) Merged_Neurons <- AddModuleScore(Merged_Neurons, features=list(genes), ctrl=15, name=paste0(sub,"_TF"))
}



Merged_Neurons <- subset(Merged_Neurons, subset = sample %in% samples_to_plot)
Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels = samples_to_plot)
Idents(Merged_Neurons) <- "sample"


# ---- Boxplot function ----
bp <- function(f){
  SCpubr::do_BoxPlot(Merged_Neurons, feature=f, split.by="sample") +
    scale_fill_manual(values=colors) +
    coord_cartesian(ylim=c(-1,1.5)) +
    ggtitle(f)
}

# ---- Plot all 4 module scores ----
bp("Lamp5_P28_TF1") | bp("Vip_P28_TF1") | bp("Sst_P28_TF1") | bp("Pvalb_P28_TF1")


# ---- Optional: inspect top TFs per subtype ----
###TFs printed per subtype, sorted by descending fold change (avg_log2FC).
lapply(subtypes, function(sub) { tfs <- top_TFs[top_TFs$cluster==sub, ]; tfs <- tfs[order(-tfs$avg_log2FC), ]; cat("\nTop TFs for", sub, "(", nrow(tfs), "TFs):", paste(tfs$gene, collapse=", "), "\n") })


```

##Figure 7 for Vln plots ##Figure 7f
```{r}
library(Seurat); library(SCpubr); library(MAST); library(patchwork); library(dplyr)

# ---- Samples, colors & subtypes ----
samples <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A",
            "Lamp5_P28"="#4D4D4D","Vip_P28"="#5F5F5F","Sst_P28"="#9E9E9E","Pvalb_P28"="#BDBDBD")
subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

# ---- Subset neurons ----
Merged_Neurons <- subset(Neurons, sample %in% samples); Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels=samples); Idents(Merged_Neurons) <- "sample"

# ---- Subset P28 interneurons ----
P28_Neurons <- subset(Merged_Neurons, sample %in% subtypes)

# ---- DEGs using MAST ----
DEG_P28_Neurons <- FindAllMarkers(P28_Neurons, test.use="MAST", logfc.threshold=0.2, min.pct=0.2, only.pos=TRUE)

# ---- Load TF list & subset ----
list_TFs <- read.csv("list_TFs.csv", header=TRUE, stringsAsFactors=FALSE)
DEG_P28_Neurons_TFs <- DEG_P28_Neurons[DEG_P28_Neurons$gene %in% list_TFs$TFs, ]

# ---- Filter TFs per subtype & compute score ----
top_TFs_list <- lapply(subtypes, function(cl){
  tmp <- DEG_P28_Neurons_TFs[DEG_P28_Neurons_TFs$cluster==cl &
                              DEG_P28_Neurons_TFs$pct.1>=0.5 & ### Keeps only genes expressed in at least 50% of cells within the cluster.Ensures the TF is robustly expressed in that subtype.
                              DEG_P28_Neurons_TFs$pct.2<=0.7 & ## Keeps only genes expressed in 60% or fewer cells outside the cluster., Ensures the TF is relatively specific, lower more stringent
                              DEG_P28_Neurons_TFs$avg_log2FC>=0.5, ] ## Only keeps genes with log2 fold-change ≥ 0.5,
  if(nrow(tmp)==0) return(NULL)
  tmp$score <- tmp$avg_log2FC * (tmp$pct.1 - tmp$pct.2)
  tmp[order(-tmp$score), ]
})
top_TFs <- do.call(rbind, top_TFs_list)

# ---- Add module scores ----
DefaultAssay(Merged_Neurons) <- "RNA"
for(cl in subtypes){genes <- top_TFs$gene[top_TFs$cluster==cl]; if(length(genes)>0) Merged_Neurons <- AddModuleScore(Merged_Neurons, list(genes), ctrl=15, name=paste0(cl,"_TF"))}

# ---- Boxplot function ----
bp <- function(f) SCpubr::do_BoxPlot(Merged_Neurons, feature=f, split.by="sample") +
  scale_fill_manual(values=colors) + coord_cartesian(ylim=c(-0.8,1.1)) + ggtitle(f)

# ---- Plot all 4 module scores across all 6 samples ----
bp("Lamp5_P28_TF1") | bp("Vip_P28_TF1") | bp("Sst_P28_TF1") | bp("Pvalb_P28_TF1")



library(patchwork)

library(patchwork)

library(patchwork)

library(patchwork)

y_text_size <- 16  # adjust as needed

(bp("Lamp5_P28_TF1") + theme(axis.text.x = element_blank(),
                               axis.ticks.x = element_blank(),
                               axis.title.x = element_blank(),
                               axis.line.x = element_blank(),
                               legend.position = "none",
                               axis.title.y = element_blank(),
                               plot.title = element_blank(),
                               axis.text.y = element_text(size = y_text_size))) /
(bp("Vip_P28_TF1") + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank(),
                            axis.line.x = element_blank(),
                            legend.position = "none",
                            axis.title.y = element_blank(),
                            plot.title = element_blank(),
                            axis.text.y = element_text(size = y_text_size))) /
(bp("Sst_P28_TF1") + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank(),
                            axis.line.x = element_blank(),
                            legend.position = "none",
                            axis.title.y = element_blank(),
                            plot.title = element_blank(),
                            axis.text.y = element_text(size = y_text_size))) /
(bp("Pvalb_P28_TF1") + theme(axis.text.x = element_blank(),
                              axis.ticks.x = element_blank(),
                              axis.title.x = element_blank(),
                              axis.line.x = element_blank(),
                              legend.position = "none",
                              axis.title.y = element_blank(),
                              plot.title = element_blank(),
                              axis.text.y = element_text(size = y_text_size)))





# ---- Optional: print top TFs per subtype ----
lapply(subtypes, function(cl){tfs <- top_TFs[top_TFs$cluster==cl, ]; tfs <- tfs[order(-tfs$score), ]; cat("\nTop TFs for", cl, "(", nrow(tfs), "):", paste(tfs$gene, collapse=", "), "\n")})

```



```{r}
lapply(seq_along(subtypes), function(i)
  data.frame(
    gene = top5_TFs[[i]],
    avg_log2FC = top_TFs$avg_log2FC[top_TFs$cluster==subtypes[i]][
      order(-top_TFs$avg_log2FC[top_TFs$cluster==subtypes[i]])][1:length(top5_TFs[[i]])]
  )
)

```



```{r}
library(Seurat); library(SCpubr); library(patchwork)

samples_levels <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A", "Lamp5_P28"="#4D4D4D","Vip_P28"="#5F5F5F","Sst_P28"="#9E9E9E","Pvalb_P28"="#BDBDBD")
subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

# Top 5 TFs per subtype
top5_TFs <- lapply(subtypes, function(cl) {
  genes <- top_TFs$gene[top_TFs$cluster==cl]
  genes <- genes[order(-top_TFs$avg_log2FC[top_TFs$cluster==cl])] #### here I order based on avg_log2FC
  head(genes,35)
})

# Create violin plots safely
plots <- lapply(seq_along(subtypes), function(i) {
  genes <- top5_TFs[[i]]
  if(length(genes)==0) return(NULL)
  VlnPlot(Merged_Neurons, features=genes, group.by="sample", fill.by="ident",
          stack=TRUE, flip=TRUE, pt.size=0, cols=colors) +
    NoLegend() + ggtitle(subtypes[i]) +
    theme(panel.grid=element_blank(), panel.border=element_rect(color="black", fill=NA), text=element_text(size=14))
})

wrap_plots(plots[!sapply(plots,is.null)], ncol=4)


library(Seurat); library(SCpubr); library(patchwork)

# ---- Flatten top5_TFs into one vector, keep track of cluster (optional for title) ----
all_top5 <- unlist(top5_TFs)
names(all_top5) <- rep(subtypes, sapply(top5_TFs,length))

# ---- Violin plot of all top TFs in one stacked plot ----
VlnPlot(Merged_Neurons, features=all_top5, group.by="sample", fill.by="ident",
        stack=TRUE, flip=TRUE, pt.size=0, cols=colors) +
  NoLegend() +
  ggtitle("Top 5 TFs per Subtype") +
  ylab("Gene Name") +
  theme(panel.grid=element_blank(),
        panel.border=element_rect(color="black", fill=NA),
        text=element_text(size=14))

```








###playing around with filters
```{r}
library(Seurat); library(SCpubr); library(MAST); library(patchwork); library(dplyr)

# ---- Samples, colors & subtypes ----
samples <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A",
            "Lamp5_P28"="#4D4D4D","Vip_P28"="#5F5F5F","Sst_P28"="#9E9E9E","Pvalb_P28"="#BDBDBD")
subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

# ---- Subset neurons ----
Merged_Neurons <- subset(Neurons, sample %in% samples); Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels=samples); Idents(Merged_Neurons) <- "sample"

# ---- Subset P28 interneurons ----
P28_Neurons <- subset(Merged_Neurons, sample %in% subtypes)

# ---- DEGs using MAST ----
DEG_P28_Neurons <- FindAllMarkers(P28_Neurons, test.use="MAST", logfc.threshold=0.2, min.pct=0.2, only.pos=TRUE)

# ---- Load TF list & subset ----
list_TFs <- read.csv("list_TFs.csv", header=TRUE, stringsAsFactors=FALSE)
DEG_P28_Neurons_TFs <- DEG_P28_Neurons[DEG_P28_Neurons$gene %in% list_TFs$TFs, ]

# ---- Filter TFs per subtype & compute score ----
top_TFs_list <- lapply(subtypes, function(cl){
  tmp <- DEG_P28_Neurons_TFs[DEG_P28_Neurons_TFs$cluster==cl &
                              DEG_P28_Neurons_TFs$pct.1>=0.3 & ### Keeps only genes expressed in at least 50% of cells within the cluster.Ensures the TF is robustly expressed in that subtype.
                              DEG_P28_Neurons_TFs$pct.2<=0.8 & ## Keeps only genes expressed in 60% or fewer cells outside the cluster., Ensures the TF is relatively specific, lower more stringent
                              DEG_P28_Neurons_TFs$avg_log2FC>=0.5, ] ## Only keeps genes with log2 fold-change ≥ 0.5,
  if(nrow(tmp)==0) return(NULL)
  tmp$score <- tmp$avg_log2FC * (tmp$pct.1 - tmp$pct.2)
  tmp[order(-tmp$score), ]
})
top_TFs <- do.call(rbind, top_TFs_list)

# ---- Add module scores ----
DefaultAssay(Merged_Neurons) <- "RNA"
for(cl in subtypes){genes <- top_TFs$gene[top_TFs$cluster==cl]; if(length(genes)>0) Merged_Neurons <- AddModuleScore(Merged_Neurons, list(genes), ctrl=15, name=paste0(cl,"_TF"))}

# ---- Boxplot function ----
bp <- function(f) SCpubr::do_BoxPlot(Merged_Neurons, feature=f, split.by="sample") +
  scale_fill_manual(values=colors) + coord_cartesian(ylim=c(-0.3,0.4)) + ggtitle(f)

# ---- Plot all 4 module scores across all 6 samples ----
bp("Lamp5_P28_TF1") | bp("Vip_P28_TF1") | bp("Sst_P28_TF1") | bp("Pvalb_P28_TF1")



library(patchwork)

library(patchwork)

library(patchwork)

library(patchwork)

y_text_size <- 16  # adjust as needed

(bp("Lamp5_P28_TF1") + theme(axis.text.x = element_blank(),
                               axis.ticks.x = element_blank(),
                               axis.title.x = element_blank(),
                               axis.line.x = element_blank(),
                               legend.position = "none",
                               axis.title.y = element_blank(),
                               plot.title = element_blank(),
                               axis.text.y = element_text(size = y_text_size))) /
(bp("Vip_P28_TF1") + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank(),
                            axis.line.x = element_blank(),
                            legend.position = "none",
                            axis.title.y = element_blank(),
                            plot.title = element_blank(),
                            axis.text.y = element_text(size = y_text_size))) /
(bp("Sst_P28_TF1") + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank(),
                            axis.line.x = element_blank(),
                            legend.position = "none",
                            axis.title.y = element_blank(),
                            plot.title = element_blank(),
                            axis.text.y = element_text(size = y_text_size))) /
(bp("Pvalb_P28_TF1") + theme(axis.text.x = element_blank(),
                              axis.ticks.x = element_blank(),
                              axis.title.x = element_blank(),
                              axis.line.x = element_blank(),
                              legend.position = "none",
                              axis.title.y = element_blank(),
                              plot.title = element_blank(),
                              axis.text.y = element_text(size = y_text_size)))





# ---- Optional: print top TFs per subtype ----
lapply(subtypes, function(cl){tfs <- top_TFs[top_TFs$cluster==cl, ]; tfs <- tfs[order(-tfs$score), ]; cat("\nTop TFs for", cl, "(", nrow(tfs), "):", paste(tfs$gene, collapse=", "), "\n")})
```

### For Figure 7d (see prepocessing 13_Figure_4_5_6)
```{r}
library(Seurat); library(ggplot2)

cols <- c("Ascl1SA6Bcl2_2wpi" = "#9400D3","Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A","Ascl1SA6Dlx2_2wpi" = "#32CD32","Other"="lightgrey")
rv <- c("Ascl1SA6Bcl2_2wpi", "Ascl1SA6Dlx2_2wpi", "Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi")

cols <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A","Other"="lightgrey")
rv <- c( "Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi")


A2Ns_900$highlight <- factor(ifelse(as.character(A2Ns_900$RV) %in% rv, as.character(A2Ns_900$RV),"Other"),
                             levels = names(cols))
DimPlot(A2Ns_900, group.by="highlight", cols=cols, pt.size=0.5) + Blank_Theme() + scale_x_reverse() + scale_y_reverse()

```



##Figure 7f and  S14b
```{r}
library(Seurat)
library(SCpubr)
library(MAST)
library(patchwork)
library(dplyr)

# ---- Samples, colors & subtypes ----
samples <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi",
             "Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD",
            "Ascl1SA6Dlx2_4wpi"="#00FA9A",
            "Lamp5_P28"="#4D4D4D",
            "Vip_P28"="#5F5F5F",
            "Sst_P28"="#9E9E9E",
            "Pvalb_P28"="#BDBDBD")

subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

# ---- Subset neurons ----
Merged_Neurons <- subset(Neurons, sample %in% samples)
Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels = samples)
Idents(Merged_Neurons) <- "sample"

# ---- Subset P28 interneurons ----
P28_Neurons <- subset(Merged_Neurons, sample %in% subtypes)

# ---- DEGs using MAST (no downstream filtering) ----
DEG_P28_Neurons <- FindAllMarkers(
  P28_Neurons,
  test.use = "MAST",
  logfc.threshold = 0.2,
  min.pct = 0.2,
  only.pos = TRUE
)

# ---- Load TF list & subset ----
list_TFs <- read.csv("list_TFs.csv", stringsAsFactors = FALSE)
DEG_P28_Neurons_TFs <- DEG_P28_Neurons[
  DEG_P28_Neurons$gene %in% list_TFs$TFs, ]

# ---- Rank TFs per subtype (NO FILTERING) ----
top_TFs_list <- lapply(subtypes, function(cl) {
  tmp <- DEG_P28_Neurons_TFs[DEG_P28_Neurons_TFs$cluster == cl, ]
  if (nrow(tmp) == 0) return(NULL)

  tmp$score <- tmp$avg_log2FC * (tmp$pct.1 - tmp$pct.2)
  tmp[order(-tmp$score), ]
})

top_TFs <- do.call(rbind, top_TFs_list)

# ---- Add module scores ----
DefaultAssay(Merged_Neurons) <- "RNA"

for (cl in subtypes) {
  genes <- top_TFs$gene[top_TFs$cluster == cl]
  if (length(genes) > 0) {
    Merged_Neurons <- AddModuleScore(
      Merged_Neurons,
      features = list(genes),
      ctrl = 15,
      name = paste0(cl, "_TF")
    )
  }
}

# ---- Boxplot function ----
bp <- function(f)
  SCpubr::do_BoxPlot(Merged_Neurons, feature = f, split.by = "sample") +
  scale_fill_manual(values = colors) +
  coord_cartesian(ylim = c(-0.5, 1.1)) +
  ggtitle(f)

# ---- Plot module scores ----
y_text_size <- 16

(bp("Lamp5_P28_TF1") + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank(),
                            axis.line.x = element_blank(),
                            legend.position = "none",
                            axis.title.y = element_blank(),
                            plot.title = element_blank(),
                            axis.text.y = element_text(size = y_text_size))) /
(bp("Vip_P28_TF1") + theme(axis.text.x = element_blank(),
                          axis.ticks.x = element_blank(),
                          axis.title.x = element_blank(),
                          axis.line.x = element_blank(),
                          legend.position = "none",
                          axis.title.y = element_blank(),
                          plot.title = element_blank(),
                          axis.text.y = element_text(size = y_text_size))) /
(bp("Sst_P28_TF1") + theme(axis.text.x = element_blank(),
                          axis.ticks.x = element_blank(),
                          axis.title.x = element_blank(),
                          axis.line.x = element_blank(),
                          legend.position = "none",
                          axis.title.y = element_blank(),
                          plot.title = element_blank(),
                          axis.text.y = element_text(size = y_text_size))) /
(bp("Pvalb_P28_TF1") + theme(axis.text.x = element_blank(),
                            axis.ticks.x = element_blank(),
                            axis.title.x = element_blank(),
                            axis.line.x = element_blank(),
                            legend.position = "none",
                            axis.title.y = element_blank(),
                            plot.title = element_blank(),
                            axis.text.y = element_text(size = y_text_size)))

# ---- Print ranked TFs per subtype ----
lapply(subtypes, function(cl) {
  tfs <- top_TFs[top_TFs$cluster == cl, ]
  cat("\nTFs for", cl, "(", nrow(tfs), "):",
      paste(tfs$gene, collapse = ", "), "\n")
})

```

```{r}
library(Seurat); library(SCpubr); library(patchwork)

samples_levels <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi","Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")
colors <- c("Ascl1SA6Bcl2_4wpi"="#6A5ACD","Ascl1SA6Dlx2_4wpi"="#00FA9A", "Lamp5_P28"="#4D4D4D","Vip_P28"="#5F5F5F","Sst_P28"="#9E9E9E","Pvalb_P28"="#BDBDBD")
subtypes <- c("Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

# Top 5 TFs per subtype
top5_TFs <- lapply(subtypes, function(cl) {
  genes <- top_TFs$gene[top_TFs$cluster==cl]
  genes <- genes[order(-top_TFs$avg_log2FC[top_TFs$cluster==cl])] #### here I order based on avg_log2FC
  head(genes,35)
})

top5_TFs <- lapply(subtypes, function(cl) {
  tmp <- top_TFs[top_TFs$cluster == cl, ]
  tmp <- tmp[order(-tmp$avg_log2FC), ]
  head(tmp$gene, 35)
})


# Create violin plots safely
plots <- lapply(seq_along(subtypes), function(i) {
  genes <- top5_TFs[[i]]
  if(length(genes)==0) return(NULL)
  VlnPlot(Merged_Neurons, features=genes, group.by="sample", fill.by="ident",
          stack=TRUE, flip=TRUE, pt.size=0, cols=colors) +
    NoLegend() + ggtitle(subtypes[i]) +
    theme(panel.grid=element_blank(), panel.border=element_rect(color="black", fill=NA), text=element_text(size=14))
})

wrap_plots(plots[!sapply(plots,is.null)], ncol=4)


library(Seurat); library(SCpubr); library(patchwork)

# ---- Flatten top5_TFs into one vector, keep track of cluster (optional for title) ----
all_top5 <- unlist(top5_TFs)
names(all_top5) <- rep(subtypes, sapply(top5_TFs,length))

# ---- Violin plot of all top TFs in one stacked plot ----
VlnPlot(Merged_Neurons, features=all_top5, group.by="sample", fill.by="ident",
        stack=TRUE, flip=TRUE, pt.size=0, cols=colors) +
  NoLegend() +
  ggtitle("Top 5 TFs per Subtype") +
  ylab("Gene Name") +
  theme(panel.grid=element_blank(),
        panel.border=element_rect(color="black", fill=NA),
        text=element_text(size=14))

```




```{r}
library(dplyr)
library(pheatmap)

samples_all <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi",
                 "Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

features_to_test <- c("Lamp5_P28_TF1",
                      "Vip_P28_TF1",
                      "Sst_P28_TF1",
                      "Pvalb_P28_TF1")

Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels = samples_all)

# Aggregate mean module scores per sample
agg_scores <- Merged_Neurons@meta.data %>%
    dplyr::select(sample, all_of(features_to_test)) %>%
    group_by(sample) %>%
    summarise(across(all_of(features_to_test), mean, na.rm = TRUE)) %>%
    column_to_rownames("sample")

# Correlation matrix (samples x samples)
cor_matrix <- cor(t(agg_scores), method = "pearson")

# Define color palette
cols <- colorRampPalette(c("white", "magenta", "#B000B0","#352A38"))(50)

# Function to map correlation value to text color
cell_fun <- function(j, i, x, y, width, height, fill) {
    # If background is dark magenta (top ~20% of palette), set text to white
    col_index <- round(cor_matrix[i, j] * (length(cols) - 1)) + 1
    text_col <- ifelse(col_index > length(cols) * 0.7, "white", "black")
    grid::grid.text(sprintf("%.2f", cor_matrix[i, j]), x, y, gp = grid::gpar(col = text_col, fontsize = 14))
}

# Plot heatmap
pheatmap(cor_matrix,
         main = "Sample Correlation (TF Module Scores)",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         color = cols,
         display_numbers = FALSE,  # numbers handled in cell_fun
         cell_fun = cell_fun)


# Save correlation matrix
write.csv(cor_matrix, "~/Desktop/samples_TF_module_score_correlation.csv")
```

### FInale for Figure with White text on dark field ##Figure 7g
```{r}
library(dplyr)
library(pheatmap)
library(grid)

samples_all <- c("Ascl1SA6Bcl2_4wpi","Ascl1SA6Dlx2_4wpi",
                 "Lamp5_P28","Vip_P28","Sst_P28","Pvalb_P28")

features_to_test <- c("Lamp5_P28_TF1",
                      "Vip_P28_TF1",
                      "Sst_P28_TF1",
                      "Pvalb_P28_TF1")

Merged_Neurons$sample <- factor(Merged_Neurons$sample, levels = samples_all)

# Aggregate mean module scores per sample
agg_scores <- Merged_Neurons@meta.data %>%
    dplyr::select(sample, all_of(features_to_test)) %>%
    group_by(sample) %>%
    summarise(across(all_of(features_to_test), mean, na.rm = TRUE)) %>%
    column_to_rownames("sample")

# Correlation matrix
cor_matrix <- cor(t(agg_scores), method = "pearson")

# Color palette
cols <- colorRampPalette(c("white", "#E000E0", "#B000B0", "#352A38"))(50)

# Cell function with direct threshold
cell_fun <- function(j, i, x, y, width, height, fill) {
    val <- cor_matrix[i, j]
    text_col <- ifelse(val > 0.55, "white", "black")
    grid.text(sprintf("%.2f", val),
              x, y,
              gp = gpar(col = text_col, fontsize = 14))
}

pheatmap(cor_matrix,
         main = "Sample Correlation (TF Module Scores)",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         color = cols,
         display_numbers = FALSE,
         cell_fun = cell_fun)

```



##Figure S14 a
```{r}
plots <- lapply(subtypes, function(cl) {

  # Subset cluster
  tmp <- top_TFs[top_TFs$cluster == cl, ]
  if (nrow(tmp) == 0) return(NULL)

  # Rank score: effect size × expression frequency
  tmp$rank_score <- tmp$avg_log2FC * tmp$pct.1

  # Order descending
  tmp <- tmp[order(-tmp$rank_score), ]

  # Select top genes
  genes <- tmp$gene[1:min(35, nrow(tmp))]

  VlnPlot(Merged_Neurons,
          features = genes,
          group.by = "sample",
          fill.by = "ident",
          stack = TRUE,
          flip = TRUE,
          pt.size = 0,
          cols = colors) +
    NoLegend() +
    ggtitle(paste0(cl, " (ranked by log2FC × pct.1)")) +
    theme(panel.grid = element_blank(),
          panel.border = element_rect(color = "black", fill = NA),
          text = element_text(size = 14))
})

wrap_plots(plots[!sapply(plots, is.null)], ncol = 4)

```


##Figure S14 a
```{r}

plots <- lapply(subtypes, function(cl) {

  tmp <- top_TFs[top_TFs$cluster == cl, ]
  if (nrow(tmp) == 0) return(NULL)

  tmp$rank_score <- tmp$avg_log2FC * tmp$pct.1
  tmp <- tmp[order(-tmp$rank_score), ]

  genes <- tmp$gene[1:min(35, nrow(tmp))]

  VlnPlot(Merged_Neurons,
          features = genes,
          group.by = "sample",
          fill.by = "ident",
          stack = TRUE,
          flip = TRUE,
          pt.size = 0,
          cols = colors) +
    NoLegend() +
    ggtitle(paste0(cl, " (ranked by log2FC × pct.1)")) +
    theme(
      panel.grid = element_blank(),
      panel.border = element_rect(color = "black", fill = NA),
      plot.title = element_text(size = 12)   # smaller title only
    )
})


# Create final combined plot
final_plot <- wrap_plots(plots[!sapply(plots, is.null)], ncol = 4)

# Save square TIFF
ggsave("~/Desktop/Top_TFs_per_Subtype_square.tiff",
       plot = final_plot,
       width = 14,
       height = 14,
       units = "in",
       dpi = 600,
       compression = "lzw")

# Optional: also save vector PDF
ggsave("~/Desktop/Top_TFs_per_Subtype_square.pdf",
       plot = final_plot,
       width = 14,
       height = 14,
       units = "in")


```

