---
title: "Figure 4"
output: html_document
date: "2025-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library(Seurat)
library(SeuratDisk)
library(patchwork)
library(ggplot2)
library(cowplot)
library(gridExtra)
library(tidyverse)
library(RColorBrewer)
library(dittoSeq)
library(Nebulosa)
library(dplyr)
library(magrittr)
library(patchwork)
library(scCustomize)
library(qs)
library(viridis)
```


```{r install.data, eval=FALSE, include=FALSE}
mergeAll_filtered <- readRDS("/seurat_4dpi_merged_LogNorm_newID_3000_removedlowcounts.rds")
```

```{r}
mergeAll_filtered<- FindClusters(object = mergeAll_filtered, resolution = 0.05)
DimPlot(mergeAll_filtered, group.by = "seurat_clusters", shuffle = TRUE)
```

```{r Cluster names, warning = FALSE, message = FALSE}
Idents(mergeAll_filtered) <- "seurat_clusters"
mergeAll_filtered<- RenameIdents(mergeAll_filtered,
   `0` = "Astro", 
`3` = "Microglia",
`1` = "Astro",
`2` = "OPCs",
`4` = "Astro", 
    `5` = "P2N", 
    `6` = "Unknown")
mergeAll_filtered <- StashIdent(mergeAll_filtered, save.name = 'cluster')
mergeAll_filtered[["cluster"]] <- Idents(object = mergeAll_filtered)
DimPlot(mergeAll_filtered, group.by = 'cluster', pt.size = 0.5) + Blank_Theme() 
```

```{r}
mergeAll_filtered$RV <- factor(mergeAll_filtered$RV,levels=c("Control", "Ascl1WT", "Ascl1WTBcl2", "Ascl1WTSA6","Ascl1SA6Bcl2","Ascl1SA6", "Ascl1SA62", "Ascl1SA6Dlx2","SA6Dlx2", "Ascl1SA6Dlx2Bcl2", "Ascl1WTDlx2", "Dlx2"))
```

```{r Cluster names, warning = FALSE, message = FALSE}
A2N <- subset(mergeAll_filtered, subset = cluster %in% c("Astro"))
```

```{r Visualisation, warning = FALSE, message = FALSE, , fig.width=6, fig.height=4 }
print(table(A2N$RV))
Idents(A2N) <- "RV"
A2N <- subset(A2N, downsample=1500)
print(table(A2N$RV))
```

```{r}
A2N2<- NormalizeData(object = A2N)
A2N2<- FindVariableFeatures(object = A2N2)
A2N2<- ScaleData(object = A2N2)
A2N2<- RunPCA(object = A2N2)
A2N2<- FindNeighbors(object = A2N2, dims = 1:23)
A2N2<- FindClusters(object = A2N2, resolution = 1.6)
A2N2<- RunUMAP(object = A2N2, dims = 1:21)#seed.use =100,
DimPlot(A2N2, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend() & Blank_Theme() 
DimPlot(A2N2, group.by = 'seurat_clusters', pt.size = 0.5, label=TRUE) + Blank_Theme() 
DimPlot(A2N2, group.by = 'RV', pt.size = 0.5, label=FALSE) + Blank_Theme() 
DimPlot(A2N2, group.by = 'Phase', pt.size = 0.5, label=TRUE) + Blank_Theme() 
### remove low count cells 
A2N2 <- subset(A2N2, subset = seurat_clusters %in% c(27, 19, 28), invert = TRUE)
A2N2<- FindClusters(object = A2N2, resolution = 1.7)
DimPlot(A2N2, group.by = 'seurat_clusters', pt.size = 0.5, label=TRUE) + Blank_Theme() 
```

```{r install.data, eval=FALSE, include=FALSE}
A2N2 <- readRDS("4dpi_A2N2_1500.rds")
```


### Figure 4d 
```{r}
RV_cols <- c("Control"="black","Ascl1SA6Bcl2"="#9400D3","SA6Dlx2"="#32CD32"); 
DimPlot(subset(A2N2, RV %in% names(RV_cols)), group.by="RV", cols=RV_cols, reduction="umap", pt.size=0.5, label=FALSE) + Blank_Theme()
```
```{r}
RV_cols <- c("Control"="black","Ascl1WTBcl2"="turquoise","Ascl1SA6Bcl2"="#9400D3","Dlx2"="#4B7A5D","SA6Dlx2"="#32CD32"); 
DimPlot(subset(A2N2, RV %in% names(RV_cols)), group.by="RV", cols=RV_cols, reduction="umap", pt.size=0.5, label=FALSE) + Blank_Theme()
```


## preperation for Figures 
```{r Cluster names, warning = FALSE, message = FALSE}
Idents(A2N2) <- "seurat_clusters"
A2N2<- RenameIdents(A2N2,
`0` = "Adv-Neuronal-1", 
`1` = "Interm-1",
`2` = "Adv-Neuronal-1",
`3` = "Interm-1",
`4` = "E-Neuronal", 
 `5` = "E-Neuronal", 
`6` = "Interm-1",
`7` = "Astro-L", 
`8` = "Adv-Neuronal-1",
`9` = "Cycling_Astro",
`10` = "Cycling_Astro",
`11` = "Adv-Neuronal-1", 
`12` = "Astro-L",
`13` = "Astrocytes",
`14` = "Adv-Neuronal-1",
`15` = "Astrocytes", 
`16` = "Astrocytes", 
`17` = "Cycling_Astro",
 `18` = "Cycling_Astro", 
`19` = "E-Neuronal",
`20` = "E-Neuronal",
`21` = "Adv-Neuronal-1",
`22` = "E-Neuronal",
 `23` = "Cycling_Astro", 
`24` = "Adv-Neuronal-1",
`25` = "Adv-Neuronal-1", 
`26` = "Astrocytes",
`27` = "Astrocytes",
`28` = "Cycling_Astro")
A2N2 <- StashIdent(A2N2, save.name = 'cell.type')
A2N2[["cell.type"]] <- Idents(object = A2N2)
DimPlot(A2N2, group.by = 'cell.type', pt.size = 0.5, label=TRUE) + Blank_Theme() 
```




### Figure 4e
```{r}
library(Seurat)
library(ggplot2)
library(ggrepel)

A2N2_4 <- subset(A2N2, subset = RV %in% c("Control","Dlx2","Ascl1WTBcl2","Ascl1SA6Bcl2","SA6Dlx2"))
Idents(A2N2_4) <- "RV"
A2N2_4$RV <- factor(A2N2_4$RV,levels=c("Control","Ascl1WTBcl2","Ascl1SA6Bcl2","Dlx2","SA6Dlx2"))
A2N2_4 <- JoinLayers(A2N2_4)
neuronlike_cells <- subset(A2N2_4, subset = cell.type %in% c("Adv-Neuronal-1") & RV %in% c("Ascl1SA6Bcl2", "SA6Dlx2"))
Idents(neuronlike_cells) <- neuronlike_cells$RV
table(neuronlike_cells$seurat_clusters, neuronlike_cells$RV)
neuronlike_cells <- JoinLayers(neuronlike_cells)
DEG <- FindMarkers(neuronlike_cells, ident.1 = "SA6Dlx2", ident.2 = "Ascl1SA6Bcl2", logfc.threshold = 0, min.pct = 0.1)
DEG$diffexpressed <- "NO"
DEG$diffexpressed[DEG$avg_log2FC > 0.5 & DEG$p_val_adj < 0.05] <- "UP"
DEG$diffexpressed[DEG$avg_log2FC < -0.5 & DEG$p_val_adj < 0.05] <- "DOWN"
top25_up <- head(rownames(DEG[DEG$diffexpressed == "UP", ][order(DEG[DEG$diffexpressed == "UP", ]$p_val_adj), ]), 1)
top25_down <- head(rownames(DEG[DEG$diffexpressed == "DOWN", ][order(DEG[DEG$diffexpressed == "DOWN", ]$p_val_adj), ]), 1)
custom_genes <- c("Pvalb", "Dlx2", "Dlx1", "Robo1","Robo2", "Olig1","Olig2", "Dlx5", "Dlx6","Gad1", "Gad2", "Dlx1as", "Sp9", "Arx", "St18","Shisa9","Rbfox3" , "Nav2", "Syt6","Mef2d", "Sox6", "Prdx5","Ckb", "St18", "Dlx6os1", "Sox2ot", "Nrxn3", "Dlx1a", "Syt6", "Dll1")
top_genes <- unique(c(top25_up, top25_down, custom_genes))
DEG$delabel <- ifelse(rownames(DEG) %in% top_genes, rownames(DEG), NA)

ggplot(DEG, aes(avg_log2FC, -log10(p_val_adj))) +
  geom_point(aes(color = diffexpressed, alpha = diffexpressed), size = 2.5) +
  scale_color_manual(values = c(DOWN = "purple", NO = "grey", UP = "#32CD32")) +
  scale_alpha_manual(values = c(DOWN = 0.2, NO = 0.4, UP = 0.6)) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "gray") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") +
  geom_text_repel(aes(label = delabel), size = 6, color = "black", segment.color = "black", box.padding = 0.5, max.overlaps = 100) +
  labs(x = "log2 Fold Change", y = "-log10(p-adj)", title = "Volcano Plot: SA6Dlx2 vs Ascl1SA6Bcl2", color = "Diff Expr", alpha = NULL) +
  guides(alpha = "none") +
  theme_minimal(base_family = "Arial") +
  theme(panel.background = element_blank(), panel.grid = element_blank(),
        plot.background = element_rect(fill = "white", color = NA),
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black"),
        plot.title = element_text(color = "black"),
        legend.background = element_rect(fill = "white"),
        legend.text = element_text(color = "black"),
        legend.title = element_text(color = "black"))
```


### Figure 4f umap

```{r}
library(scCustomize)
library(patchwork)

A2N2_3 <- subset(A2N2, subset = RV %in% c("Control",  "Ascl1SA6Bcl2","SA6Dlx2"))
Idents(A2N2_3) <- "RV"
A2N2_3$RV <- factor(A2N2_3$RV,levels=c("Control",  "Ascl1SA6Bcl2","SA6Dlx2"))
A2N2_3 <- JoinLayers(A2N2_3)
fp <- FeaturePlot_scCustom( seurat_object = A2N2_3, features = c("Olig1","Olig2"),  colors_use = colorRampPalette(c("white","grey","#6495ED","#00008B", "black"))(100), combine = FALSE)
(fp_clean <- wrap_plots(lapply(fp, function(p) p + Blank_Theme()), ncol = 1))
```
```{r}
library(scCustomize)
library(patchwork)

A2N2_3 <- subset(A2N2, subset = RV %in% c("Control",  "Ascl1SA6Bcl2","SA6Dlx2"))
Idents(A2N2_3) <- "RV"
A2N2_3$RV <- factor(A2N2_3$RV,levels=c("Control",  "Ascl1SA6Bcl2","SA6Dlx2"))
A2N2_3 <- JoinLayers(A2N2_3)
fp <- FeaturePlot_scCustom( seurat_object = A2N2_3, features = c("Htr3a"),  colors_use = colorRampPalette(c("white","grey","#6495ED","#00008B", "black"))(100), combine = FALSE)
(fp_clean <- wrap_plots(lapply(fp, function(p) p + Blank_Theme()), ncol = 1))
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Fetch expression + RV for only cells with Htr3a
df <- FetchData(A2N2_3, vars = c("Htr3a")) %>%
  mutate(Cell = rownames(.)) %>%
  left_join(
    data.frame(Cell = colnames(A2N2_3), RV = Idents(A2N2_3)),
    by = "Cell"
  )

# Plot with 3 colors (add black for the third RV)
df %>%
  pivot_longer(cols = c(Htr3a), names_to = "Gene", values_to = "Expr") %>%
  ggplot(aes(Gene, Expr, fill = RV)) +
  geom_boxplot(position = position_dodge(0.8)) +
  scale_fill_manual(values = c( "black","#9400D3", "#32CD32")) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank()) +
  labs(y = "Expression", x = "Gene", title = "Htr3a by RV")

```





```{r}
library(scCustomize)
library(patchwork)

A2N2_WT <- subset(A2N2, subset = RV %in% c("Control","Dlx2","Ascl1WT", "Ascl1WTDlx2"))
Idents(A2N2_WT) <- "RV"
A2N2_WT$RV <- factor(A2N2_WT$RV,levels=c("Ascl1WT", "SA6Dlx2", "Ascl1SA6Dlx2Bcl2", "Ascl1WTDlx2", "Dlx2"))
A2N2_WT <- JoinLayers(A2N2_WT)
fp <- FeaturePlot_scCustom( seurat_object = A2N2_WT, features = c("Olig1","Olig2"),  colors_use = colorRampPalette(c("white","grey","#6495ED","#00008B", "black"))(100), combine = FALSE)
(fp_clean <- wrap_plots(lapply(fp, function(p) p + Blank_Theme()), ncol = 1))
```

### Slingshot lineages for 5 samples Fig 3d
```{r}
library(slingshot)
A2N2_3 <- JoinLayers(A2N2_3)
sce <- as.SingleCellExperiment(A2N2_3, assay = "RNA")
sce_cl3 <- slingshot(sce,
                     clusterLabels = "cell.type",
                     start.clus = c("Cycling_Astro"), ### cluster from unbiased latent time 
                     approx_points = 200,
                     extend = "n",
                     reducedDim = "UMAP")

slingLineages(sce_cl3)

A2N2_3$Pseudotime_1 <- sce_cl3$slingPseudotime_1
A2N2_3$Pseudotime_2 <- sce_cl3$slingPseudotime_2
A2N2_3$Pseudotime_3 <- sce_cl3$slingPseudotime_3
A2N2_3$Pseudotime_4 <- sce_cl3$slingPseudotime_4
A2N2_3$Pseudotime_5 <- sce_cl3$slingPseudotime_5
A2N2_3$Pseudotime_6 <- sce_cl3$slingPseudotime_6
A2N2_3$Pseudotime_7 <- sce_cl3$slingPseudotime_7
A2N2_3$Pseudotime_8 <- sce_cl3$slingPseudotime_8
A2N2_3$Pseudotime_9 <- sce_cl3$slingPseudotime_9
A2N2_3$Pseudotime_10 <- sce_cl3$slingPseudotime_10

library(ggpubr); library(viridis); library(purrr); library(stringr)
curves <- map(1:2, ~ slingCurves(sce_cl3)[[.x]]$s[slingCurves(sce_cl3)[[.x]]$ord, ] %>%
  as.data.frame() %>% setNames(c("umap_1", "umap_2")))
plots <- map(1:2, ~ FeaturePlot(A2N2_3, features = str_c("Pseudotime_", .x), reduction = "umap") +
  scale_colour_viridis() + Blank_Theme() +
  geom_path(data = curves[[.x]], aes(x = umap_1, y = umap_2)))
ggarrange(plotlist = plots, ncol = 2, nrow = 2)

A2N2_3$Pseudotime_avg <- rowMeans(slingPseudotime(sce_cl3)[, c(1,2)], na.rm = TRUE)
plot_pseudotime <- function(feature_name, curve_index) {
  FeaturePlot(A2N2_3, features = feature_name, reduction = "umap") +
    scale_colour_viridis() + 
    theme_void() +  # Removes background, axes, and grid
    geom_path(data = as.data.frame(slingCurves(sce_cl3)[[curve_index]]$s[slingCurves(sce_cl3)[[curve_index]]$ord,]),
              aes(x = umap_1, y = umap_2))
}
plot_pseudotime <- function(feature_name, curve_index) {
  FeaturePlot(A2N2_3, features = feature_name, reduction = "umap") +
    scale_colour_viridis() + 
    theme_void() +  # Removes all background, grid, axes, and text
    theme(
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank(), 
      panel.background = element_blank()
    ) +
    geom_path(data = as.data.frame(slingCurves(sce_cl3)[[curve_index]]$s[slingCurves(sce_cl3)[[curve_index]]$ord,]),
              aes(x = umap_1, y = umap_2))
}

FeaturePlot(A2N2_3, features = "Pseudotime_avg", reduction = "umap") +
  scale_colour_viridis() + theme_void() 
```





### Figure 4f boxplots
```{r}
library(SCpubr)
Idents(neuronlike_2) <- "RV"
library(ggplot2)
library(dplyr)
library(tidyr)

FetchData(neuronlike_2, c("Olig1","Olig2")) %>%
  mutate(RV = Idents(neuronlike_2)) %>%
  pivot_longer(cols = c(Olig1, Olig2), names_to = "Gene", values_to = "Expr") %>%
  ggplot(aes(Gene, Expr, fill = RV)) +
  geom_boxplot(position = position_dodge(0.8)) +
  scale_fill_manual(values = c("#9400D3", "#32CD32")) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank()) +   # ← removes grid lines
  labs(y = "Expression", x = "Gene", title = "Olig1 and Olig2 by RV")

```

### Figure 4g umap of Dlx2 target genes 
```{r}
library(scCustomize)
library(patchwork)

fp <- FeaturePlot_scCustom(
  seurat_object = A2N2_3, features = c("Dlx1", "Dlx6", "Sp8","Zeb2","Arx" ,"Mef2c", "Gad1", "Gad2"), colors_use = colorRampPalette(c("white","grey", "#66CDAA","#2E8B57", "black"))(100), combine = FALSE)
(fp_clean <- wrap_plots(lapply(fp, function(p) p + Blank_Theme()), ncol = 4))
```


### Figure 4h umap and boxplot of Pvalb expression 
```{r}
FeaturePlot_scCustom(subset(A2N2_3), features = "Pvalb", colors_use =colorRampPalette(c("white","grey","#800080","#2E0854","black"))(100)) & Blank_Theme()
boxplot_obj2 <- SCpubr::do_BoxPlot(sample = A2N2_3, feature = "Pvalb", split.by = "RV")
boxplot_obj2 <- boxplot_obj2 + scale_fill_manual(values = c("black", "turquoise","#9400D3","#4B7A5D", "#32CD32"))
boxplot_obj2 + theme_void() + theme(legend.position = "right")

```

```{r}
library(SCpubr)
Idents(neuronlike_2) <- "RV"
library(ggplot2)
library(dplyr)
library(tidyr)

FetchData(neuronlike_2, c("Pvalb")) %>%
  mutate(RV = Idents(neuronlike_2)) %>%
  pivot_longer(cols = c(Pvalb), names_to = "Gene", values_to = "Expr") %>%
  ggplot(aes(Gene, Expr, fill = RV)) +
  geom_boxplot(position = position_dodge(0.8)) +
  scale_fill_manual(values = c("#9400D3", "#32CD32")) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),   # still remove grid
    axis.line = element_line()      # add back x/y axis lines
  ) +
  labs(y = "Expression", x = "Gene", title = "Olig1 and Olig2 by RV")
```


```{r}
library(SCpubr)
Idents(neuronlike_2) <- "RV"
library(ggplot2)
library(dplyr)
library(tidyr)

FetchData(neuronlike_2, c("Kcnc1")) %>%
  mutate(RV = Idents(neuronlike_2)) %>%
  pivot_longer(cols = c(Kcnc1), names_to = "Gene", values_to = "Expr") %>%
  ggplot(aes(Gene, Expr, fill = RV)) +
  geom_boxplot(position = position_dodge(0.8)) +
  scale_fill_manual(values = c("#9400D3", "#32CD32")) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),   # still remove grid
    axis.line = element_line()      # add back x/y axis lines
  ) +
  labs(y = "Expression", x = "Gene", title = "Kcnc1 by RV")

VlnPlot(A2N2_3, features = c( "Kcnc1"), cols = c("black","#9400D3", "#32CD32", ncol(4 )), pt.size = 0) 
```

### Figure 4i umap and boxplot of Pvalb expression 
```{r}
A2N2_fin <- subset(A2N2, subset = RV %in% c("Control","Ascl1SA6Bcl2","SA6Dlx2","Ascl1SA6Dlx2Bcl2"))
Idents(A2N2_fin) <- "RV"
A2N2_fin$RV <- factor(A2N2_fin$RV,levels=c("Control","Ascl1SA6Bcl2","SA6Dlx2","Ascl1SA6Dlx2Bcl2"))
DimPlot(A2N2_fin,  reduction = "umap", group.by = "RV",  pt.size = 1, cols = c("black","#9400D3", "#32CD32", "#FDD835")) & Blank_Theme()
```
### Figure 4j umap of Pvalb expression 
```{r}
FeaturePlot_scCustom(subset(A2N2_fin, RV %in% c("Ascl1SA6Bcl2", "Ascl1SA6Dlx2Bcl2")), features = "Pvalb", pt.size=0.5, split.by = "RV", colors_use =colorRampPalette(c("white","grey","#800080","#2E0854","black"))(100)) & Blank_Theme()
```


### Extended Data Figure 9b 
```{r}
RV_cols <- c("Control"="black","Ascl1WT"="#1E90FF","Ascl1WTBcl2"="turquoise","Ascl1SA6"="#DB7093","Ascl1SA6Bcl2"="#9400D3","Dlx2"="#4B7A5D","Ascl1WTDlx2"="#6B8E23","SA6Dlx2"="#32CD32","Ascl1SA6Dlx2Bcl2"="#FDD835"); 
DimPlot(subset(A2N2, RV %in% names(RV_cols)), group.by="RV", cols=RV_cols, reduction="umap", pt.size=0.5, label=FALSE) + Blank_Theme()
```

### Extended Data Figure 9c 
```{r}
p1  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Control"]),  cols.highlight = "black")       + ggtitle("Control")           + NoLegend() + Blank_Theme()
p2  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Ascl1WT"]),  cols.highlight = "#1E90FF")     + ggtitle("Ascl1WT")          +  NoLegend() + Blank_Theme()
p3  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Ascl1WTBcl2"]), cols.highlight = "turquoise")   + ggtitle("Ascl1WTBcl2")       + NoLegend() + Blank_Theme()
p4  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Ascl1SA6"]), cols.highlight = "#DB7093")     + ggtitle("Ascl1SA6")          + NoLegend() + Blank_Theme()
p5  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Ascl1SA6Bcl2"]),  cols.highlight = "#9400D3")     + ggtitle("Ascl1SA6Bcl2")      + NoLegend() + Blank_Theme()
p6  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Dlx2"]), cols.highlight = "#4B7A5D")     + ggtitle("Dlx2")              + NoLegend() + Blank_Theme()
p7  <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Ascl1WTDlx2"]), cols.highlight = "#6B8E23")    + ggtitle("Ascl1WTDlx2")       + NoLegend() + Blank_Theme()
p8 <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "SA6Dlx2"]), cols.highlight = "#32CD32")     + ggtitle("SA6Dlx2")           + NoLegend() + Blank_Theme()
p9 <- DimPlot(A2N2, reduction = "umap", cells.highlight = colnames(A2N2[, A2N2$RV == "Ascl1SA6Dlx2Bcl2"]), cols.highlight = "#FDD835")     + ggtitle("Ascl1SA6Dlx2Bcl2")  + NoLegend() + Blank_Theme()
(p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8+ p9) + plot_layout(ncol = 4)
```

### Extended Data Figure 9d
```{r Cluster names, warning = FALSE, message = FALSE}
cell_type_colors <- c( "Astrocytes" = "#A0C0B0", "Cycling_Astro" = "#ACE1AF",  "Astro-L" = "#8FBC8F",  "Interm-1" = "#8F9D53", "E-Neuronal" = "#006400", "Adv-Neuronal-1" = "#2F3826")
A2N2@meta.data$cell.type.color <- cell_type_colors[A2N2@meta.data$cell.type]
A2N2$cell.type <- factor(A2N2$cell.type, levels = c("Astrocytes","Cycling_Astro", "Astro-L", "Interm-1", "E-Neuronal" ,"Adv-Neuronal-1"))
DimPlot(subset(A2N2, RV %in% c("Control","Ascl1WT","Ascl1WTBcl2","Ascl1SA6","Ascl1SA6Bcl2","Dlx2","Ascl1WTDlx2","SA6Dlx2","Ascl1SA6Dlx2Bcl2")), group.by="cell.type", cols=cell_type_colors, label=FALSE, pt.size=0.5) + NoLegend() + Blank_Theme()
FeaturePlot(subset(A2N2, RV %in% c("Control","Ascl1WT","Ascl1WTBcl2","Ascl1SA6","Ascl1SA6Bcl2","Dlx2","Ascl1WTDlx2","SA6Dlx2","Ascl1SA6Dlx2Bcl2")), features = c("Mki67","Slc1a3","Aqp4","Aldh1l1","Gfap","Olig1","Olig2","Notch1","Hes5","Sox11","Tubb3","Rbfox3","Srrm4","Dcx"), ncol = 5, reduction = "umap") & Blank_Theme()
DimPlot(subset(A2N2, RV %in% c("Control","Ascl1WT","Ascl1WTBcl2","Ascl1SA6","Ascl1SA6Bcl2","Dlx2","Ascl1WTDlx2","SA6Dlx2","Ascl1SA6Dlx2Bcl2")), group.by="Phase",  label=FALSE, pt.size=0.5) + NoLegend() + Blank_Theme()
```

### GO Terms SA6Bcl2 (down) versus SA6Bcl2 (up) for Extended Data Figure 10a 
```{r}
library(clusterProfiler)
library(org.Mm.eg.db)
library(ggplot2)
library(patchwork)

up_genes <- rownames(DEG[DEG$diffexpressed == "UP", ])
down_genes <- rownames(DEG[DEG$diffexpressed == "DOWN", ])
entrez_up <- bitr(up_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Mm.eg.db)
entrez_down <- bitr(down_genes, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Mm.eg.db)
ego_up <- enrichGO(entrez_up$ENTREZID, OrgDb=org.Mm.eg.db, ont="BP", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)
ego_down <- enrichGO(entrez_down$ENTREZID, OrgDb=org.Mm.eg.db, ont="BP", pAdjustMethod="BH", qvalueCutoff=0.05, readable=TRUE)
p1 <- dotplot(ego_up, showCategory=15, title="UP-regulated Genes", font.size=14, label_format=80) + theme_minimal()
p2 <- dotplot(ego_down, showCategory=15, title="DOWN-regulated Genes", font.size=14, label_format=80) + theme_minimal()
p1/p2
```

### Boxplots for Extended Data Figure 10b
```{r}
neuronlike_2 <- subset(A2N2, subset = cell.type %in% c("Adv-Neuronal-1") & RV %in% c("Ascl1WTBcl2","Ascl1SA6Bcl2","SA6Dlx2"))
neuronlike_2 <- JoinLayers(neuronlike_2)
library(SCpubr)
Idents(neuronlike_2) <- "RV"
library(ggplot2)
library(dplyr)
library(tidyr)
FetchData(neuronlike_2, c("Dlx2")) %>%
  mutate(RV = Idents(neuronlike_2)) %>%
  pivot_longer(cols = c(Dlx2), names_to="Gene", values_to="Expr") %>%
  ggplot(aes(Gene, Expr, fill=RV)) +
  geom_boxplot(position=position_dodge(0.8)) +
  scale_fill_manual(values=c("turquoise","#9400D3","#32CD32")) +
  theme_minimal(14) +
  labs(y="Expression", x="Gene", title="Dlx2 by RV")
```
### Boxplots for Extended Data Figure 10b
```{r}
A2N2_Dlx2 <- subset(A2N2, subset = RV %in% c("Control","Dlx2","Ascl1SA6Bcl2", "SA6Dlx2"))
A2N2_Dlx2$RV <- factor(A2N2_Dlx2$RV,levels=c("Control","Dlx2","Ascl1SA6Bcl2", "SA6Dlx2"))

A2N2_Dlx2 <- JoinLayers(A2N2_Dlx2)
library(SCpubr)
Idents(A2N2_Dlx2) <- "RV"
library(ggplot2)
library(dplyr)
library(tidyr)
FetchData(A2N2_Dlx2, c("Ascl1")) %>%
  mutate(RV = Idents(A2N2_Dlx2)) %>%
  pivot_longer(cols = c(Ascl1), names_to="Gene", values_to="Expr") %>%
  ggplot(aes(Gene, Expr, fill=RV)) +
  geom_boxplot(position=position_dodge(0.8)) +
  scale_fill_manual(values=c("black","#4B7A5D","#9400D3","#32CD32")) +
  theme_minimal(14) +
  labs(y="Expression", x="Gene", title="Dlx2 by RV")
```


### umap of Dlx2 target genes for Extended Data Figure 10c
```{r}
library(scCustomize)
library(patchwork)

fp <- FeaturePlot_scCustom(
  seurat_object = A2N2_4, features = c( "Ascl1","Dlx2", "Dlx6","Sp9","Maf","Mafb", "Satb1", "St18"), colors_use = colorRampPalette(c("white","grey", "#66CDAA","#2E8B57", "black"))(100), combine = FALSE)
(fp_clean <- wrap_plots(lapply(fp, function(p) p + Blank_Theme()), ncol = 5))
```

```{r}
FetchData(neuronlike_2, c("Kcnc1")) %>%
    mutate(RV = Idents(neuronlike_2)) %>%
    pivot_longer(cols = c(Kcnc1), names_to = "Gene", values_to = "Expr") %>%
    ggplot(aes(Gene, Expr, fill = RV)) +
    geom_boxplot(position = position_dodge(0.8)) +
    scale_fill_manual(values = c("#9400D3", "#32CD32")) +
    theme_minimal(base_size = 14) +
    theme(
        panel.grid = element_blank(),   # still remove grid
        axis.line = element_line()      # add back x/y axis lines
    ) +
    labs(y = "Expression", x = "Gene", title = "Kcnc1 by RV")
```




